---
title: "Basics"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
library(reticulate)

#  md_document:
#    variant: markdown_github
#
#
#<div class = "row">
#<div class = "col-md-6">
#<div class = "python">
#  
#
#</div>
#</div>
#<div class = "col-md-6">
#<div class = "r">  
#  
#  
#</div>
#</div>
#</div>


#<style>
#div.blue pre { background-color:lightblue; }
#div.blue pre.r { background-color:blue; }
#</style>
#
```

<style>
div.python pre { background-color: #e9f7f5
; }
</style>

<style>
div.r pre { background-color: #fff5fd; }
</style>


<p>&nbsp;</p>

This capter introduces the basics of R and Python. The goal of this chapter is to discover the different types of object that we can deal with for each language, as well as understanding the logic of control flows and function writing.
The chapter ends with 3 Exercices from Euler Project. By forming a group of 2, solve at least one exercice in the language of your choice and try to compare different approaches.


<p>&nbsp;</p>



# R/Python

Both data science languages are great for business analysis. They can both be used for similar purpose when viewed from a purely machine learning perspective. Both have packages or libraries dedicated to wrangling, preprocessing, and applying machine learning to data. Both are great choices for reproducibile research. Where things become really interesting is their differences, as well as combining languages.


## R Strengths

The S language was a precursor to R developed by John Chambers (statistician) at Bell Labs in 1976 as a programming language designed to implement statistics. The R statistical programming language was developed at the University of Auckland, New Zealand, to increase S capacities. S and R developers were not software engineers or computer scientists, they were researchers that developed tools in order carry out research and communicate results more efficiently. 

R is a language with roots in statistics, data analysis, data exploration, and data visualization. It has utilities for reporting and communicate including RMarkdown and Shiny. 

R is expanding rapidly with the emergence of the tidyverse (tidyverse.org), a set of tools with a common programming-interface that use functional verbs (functions like mutate() and summarize()) to perform intuitive operations connected by the pipe (%>%). The tidyverse is very advantageous because it makes data mining very efficient. Iterating through your exploratory analysis is as simple as writing a paragraph describing what you intend to do to the data.

## Python Strengths

The Python language is a general-purpose programming language created by Guido van Rossum (Computer Scientist) in 1991. The language was developed to be readable and to cover many programming paradigms. One of Python's greatest strengths is its flexibility which allows to handle web frameworks, data base connectivity, networking, web scraping, text and image processing, many of those features are very useful to various tasks in machine learning.

In essence, the foundations of Python are in computer science and mathematics. With over 100,000 open source libraries, Python has the largest ecosystem of any programming language, making it a unique choice for those who want high flexibility.

Python has great data science libraries including Scikit Learn, the most popular machine learning library which is easy to pick up, includes support for pipelines to simplify the machine learning workflow, and has almost all of the algorithms one needs in one place. Another very common library is TensorFlow, developed by software engineers at Google to perform deep learning, commonly used for image recognition and natural language processing tasks (NLTK is also very useful for NLP). Facebook also have his own deep learning framework named PyTorch which is a concurrent of Tenserflow and Keras (which is desinged for efficiently building neural networks). 

 

- R is selected for exploration due to the tidyverse and datatable readability and efficiency.

- Python is preferred for machine learning because of Scikit Learn, TenserFlow and Keras, PyTorch machine learning pipeline capability.

- R is favored for communication because of the advanced reporting utilities including RMarkdown and Shiny (interactive web apps) and ggplot2 for visualization.

# Operations

## Arithmetic Operations
Python and R work as classical calculator, using "+", "-", "*" and "/" we can do arithmetic operations in both languages.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ope}
# Python 

1+2
1-2
1/2
1*2

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-ope, collapse = TRUE}
# R

1+2
1-2
1/2
1*2
```

</div>
</div>
</div>

<p>&nbsp;</p>

We can also apply exponantiation, Modulo and floor division easily in both language. Note that in R we can write the exponantiation "^". It is used for exclusive-or (XOR) in Python.


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ope2,echo = T,collapse = TRUE}
# Python 

2**8 # exponantiation
2^8 == 2**8 # False
8%3 # modulo
8//3 # floor division
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-ope2,echo=T,collapse = TRUE}
# R

2**8 # exponantiation
2^8 == 2**8 # TRUE
8%%3 # modulo
8%/%3 # floor division

```

</div>
</div>
</div>
Operator (R) | Operator (Python) |	Description
-|:-:|-:
+ | + |	Addition
– | – |	Subtraction
* | * |	Multiplication
/ | / |	Division
^ / ** | ** |	Exponent
%% | % |	Modulo
%/% | // |	floor Division


<p>&nbsp;</p>
##  Comparison Operators

In order to compare some values, we can use comparison operators to investigate if a given value is equal to, not equal to, greater than, etc.

Theses operators are the same in both language.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python comp,echo = T,collapse = TRUE}
# Python 

2==8
2!=8
2<8
2>8
2<=8

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-cromp,echo=T,collapse = TRUE}
# R

2==8
2!=8
2<8
2>8
2<=8
```
</div>
</div>
</div>

Operator (R/Python) |	Description
-|-:
\< 	| Less than
\> 	| Greater than
\<= 	| Less than or equal to
\>= 	| Greater than or equal to
== 	| Equal to
!= 	| Not equal to

<p>&nbsp;</p>

## Logical operators


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python logope,echo = T,collapse = TRUE}
# Python 
x = [True,True]
y = [True,False]

not x[0]
x and y
x or y
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-logope,echo=T,collapse = TRUE}
# R

x = c(TRUE,TRUE)
y = c(TRUE,FALSE)

!x[1]
x & y
x && y
x | y
x || y
```
</div>
</div>
</div>

Operator (R)| Operator (Python)|	Description
-|:-:|-:
! | not |	Logical NOT
& | and |	Element-wise logical AND
&& |  |Logical AND
\| |or|	Element-wise logical OR
\|\|  | |	Logical OR


<p>&nbsp;</p>
## Membership operators

'In' and 'not in' are membership operators in Python. They are used to test whether a value or variable is found in a sequence. In R '%in%' allows you to use the python in, but you have to be aware that they may not react the same way, the reason is that list of character can be directly decomposed in a character list in python, which is not the case in R.
See how we can check that 'Hello' is in 'Hello world' using just in with python. In R we look if the string exactly match one element.


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python relope,echo = T,collapse = TRUE}
# Python 
x = 'Hello world'
y = {1:'a',2:'b'}

print('world' in x)

print(1 in y)

print('a' in y)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-relope,echo=T,collapse = TRUE}
# R

x = 'Hello World'

print('Hello' %in% x)
stringr::str_detect(x,'Hello')
```
</div>
</div>
</div>



<p>&nbsp;</p>

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python relope2,echo = T,collapse = TRUE}
# Python 
x = ['Hello','World']

print('Hello' in x)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-relope2,echo=T,collapse = TRUE}
# R

x = c('Hello','World')

print('Hello' %in% x)
stringr::str_detect('Hello',x)
```
</div>
</div>
</div>


<p>&nbsp;</p>

# Objects & Variables


In Python and R, we do not need to declare a variable before assigning a value to this variable. we can think of a variable as a name to refers to an object. There is a difference between Python and R on the way the computer is storing objects and variables.


<p>&nbsp;</p>

## Some differences between R and Python

While Python is much appreciated for being a versatile language with an easy-to-understand syntax, R's functionalities are developed by statisticians, thus giving it field-specific advantages.

In R, a variable and an object are the same things, they refer to the same entities. If we assign to a new variable a variable that already exist, it will refers to two different objects.

Things in Python are more conventional, but since most of you learned R before Python it is important to notice that variables refers to an objet but they are still two entities.
It means that two variables can refer to a single object. If two variables are pointing the same object, the modification made with one variable will be also available when using the other variable.


We mostly should never use for loops in R - they're very slow because they execute a function call with every iteration. One should vectorize and use the apply family of functions instead (Weird, I know..). Vectorization is king in R if we want fast code. Assuming we vectorize both our R and Python codes (and other factors), we should probably get the same order of magnitude in speed. For data larger than memory (we can specify the limit), R starts to become a bad choice. 



<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python ov,echo = T,collapse = TRUE}
# Python 
import numpy as np
only_nb = [1,2,3,4]
np.std(only_nb)
np.std(only_nb,ddof=1)
with_none = [1,2,3,None]
np.std(with_none)
with_char = [1,2,'3',4]
np.std(with_char)

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-ov,echo=T,collapse = TRUE}
# R
only_nb <- c(1,2,3,4)
sd(only_nb)
with_na <- c(1,2,3,NA)
sd(with_na)
with_char <- c(1,2,'3',4)
sd(with_char)

with_char <- c(1,2,'a',4)
sd(with_char)

```
</div>
</div>

<p>&nbsp;</p>

- See how R does not show any error and directly remove NA to compute the standard deviation and also converts string to numeric if possible. If converting the string to numeric is not feasible, R finally return a NA. This is not an error message but a warning message.  
We need to write our code carefully since R can do calculation for a while and just return NA without stopping the process which is not the case in Python. 

- see here that R calulates the standard deviation with N - 1 as the denominator, and numpy with N. To get the same result we need to use 'ddof=1' as argument in std.
</div>


<p>&nbsp;</p>
## Textual and numerical variables

In R textual data are called 'character', in Python it is a string, abbreviate as 'str'. " or ' are used when we set a textual variable in both language, but we can also set a variable to textual variable if needed.

Numerical variables a decomposed in three types. In Python we have integer, float and complex. In R float is called 'double', reffering to double precision floating point. Without entering into details, it means that the number have a precision up to 15 decimals. Float can refers to a precision of 7 decimals, however in Python the data-type 'float' has a precision of 15 numbers. They are ways to get more precisions using Numpy for example. 

Finaly we can check easily the data-type of the variable that we are dealing by asking the language with the command 'type' in Python, or 'typeof' in R.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python tnv,echo = T,collapse = TRUE}
# Python 

import sys 
sys.float_info.dig # number of decimals

a = 1
type(a)

b = 1.1
type(b)

c = 1.1+2j
type(c)

d = 'd'
type(d)

# change the type
e = 2
f = str(e)
f
float(f)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-tnv,echo=T,collapse = TRUE}
# R

a = 1
typeof(a)
a_int = 1L
typeof(a_int)

b = 1.1
typeof(b)

c = 1.1+2i
typeof(c)

d = 'd'
typeof(d)

# change the type
e = 2
f = as.character(e)
f
as.numeric(f)
```
</div>
</div>
</div>

<p>&nbsp;</p>

## Other Data-Type (R)

There is a lot of datatypes in both language that we will discover along this course, for now, we will focus only on basic data-type that are directly available without using anyother package or library,in R common data-type are Vector, Matrix, List, Dataframe.

<p>&nbsp;</p>

### Vectors and Matrices

<div class = "r">

<div class = "row">
<div class = "col-md-6">

```{r r-vm}
# R

vec = c(1, 2, 3)
length(which(vec==2)) # count elements of the list which are exactly equal to 2

vec = sort(vec,
           decreasing = TRUE)
vec
# access the element of a list
vec[1]
vec[which(vec==3)]
vec[2:length(vec)]
vec[length(vec):2]
vec[-1] # throwaway value


```

</div>
<div class = "col-md-6">

```{r r-vm2}
# R

mat = matrix(c(1, 2, 3, 4, 5, 6),3,2)
dim(mat)
mat

length(which(mat==2)) # count elements of the list which are exactly equal to 2

vec = sort(mat,
           decreasing = TRUE)
vec
# access the element of a list
mat[1] # first element
which(mat==3,
      arr.ind = TRUE) # get row and columns for each value that match

mat[2:length(mat)] # gives us a vector
mat[2:dim(mat)[1],]
mat[length(mat):1]

mat[1,-1] # throwaway values


```

</div>
</div>

```{r r-vm3}
# modify

b = rep(0,3)
vec = c(vec,b)
vec
mat_2 = rbind(vec,vec)
mat_2 
mat_2 = cbind(mat_2,mat_2)
mat_2

t_mat = t(mat) # transposition
t_mat[2,3] = as.character(t_mat[2,3]) # all values turn to character


i_mat = diag(1, 3)
i_mat


```


</div>

<p>&nbsp;</p>

### Data frame

A Data frame in R is a matrix which can store different types of data, characters, numerics or factors. Contrary to the matrix, it does not have a fixed dimension which means that we can create a column very easily using the operator '$'.

<div class = "r">

```{r r-df,echo=T,collapse = TRUE}
# R

df = data.frame(chr = letters[1:5], # letters is a vector of letter provided directly in R
                num = seq(1,5,1),
                fac = factor(c(rep('a',3),rep('b',2)))) 
class(df$chr) # automaticaly converted to factor, which can be very inneficient !
df$chr = as.character(df$chr)
class(df$chr)
colnames(df)
df
# access the element of a variable in a df
df$chr[3]
which(df==3,
      arr.ind = TRUE) 

df[1:2] # gives us a df with the first two columns !
df[3:dim(df)[1],]

df[1,-1] # throwaway values

```
```{r r-df2,echo = T,collapse = TRUE}
# modify

b = rep(0,3)
df = rbind(df,b) # introduce a new row in the data frame, note that we get a NA in the factor colmun because '0' is not a level
df$new_col = rep(2,6) # we can introduce new colmuns by specifiying the name and set a value


df
t_df = t(df) # transposition convert the df into a matrix
class(t_df)
```
</div>

<p>&nbsp;</p>

### Arrays

Arrays is a generalisation of the matrix, meaning that we can have a matrix with more than two dimensions. We can modify them as we modify a matrix.

<div class = "r">

```{r r-ar,echo=T,collapse = TRUE}
# R

arr = array(seq(1,12,1), dim = c(2,2,3)) # create a matrix with 3 dimension
arr
# access the element of a dictionary
arr[1,,]
arr[,1,]
arr[,,1]
```
</div>


<p>&nbsp;</p>

### Lists 

A list permits us to assemble an assortment of disconnected objects behind one object, it is also possible to store objects which can be of different types.


<div class = "r">

```{r r-lst,echo=T,collapse = TRUE}
# R

lst = list(df = df,
           mat = mat,
           arr = arr) # create a matrix with 3 dimension

lst
lst[[1]] # first element of the list, here it is the df
lst[[1]][1,] # first row of first element of the list
```
</div>



<p>&nbsp;</p>

## Other Data-Type (Python)



In Python the most common are List, tuple, sets, Dictionnary.
Note that for the moment, for modifiying a variable we always use '='. Keep in mind that using the '.' can modify the object behind the variable in Python.

<p>&nbsp;</p>

### Lists 

Keep in mind that Lists are mutable, as discussed in the Section "Difference between R and Python"

<div class = "python">
```{python lst,echo = T,collapse = TRUE}
# Python

a = [1, 2, 3]
a.count(2) # count elements of the list which are exactly equal to 2

a.sort(reverse = True)
a
# access the element of a list
a[0]
a.index(3)
a[1:]
a[:1]
a[0:-1]
a[:]

```
```{python lst2,echo = T,collapse = TRUE}
# modify

b = [0, 0, 0]
list(zip(a,b)) # zip will pairs the ellements, it works also with more than 2 element ex: zip(a,b,c)
a.append(b)
a
a[4:5] = ['a','b']
a
a.extend([4, 5, 6])
a
a += [7,8] # works as extend
a
a.insert(2,[1,2])
a
a.remove(b)
a

a = a*2 # replicate the list n times
len(a) # number of elements in the list

```

```{python lst3,echo = T,collapse = TRUE}
# mutable

a = [1,2,3]
b = a
b[0] = 12
a

```
</div>



<p>&nbsp;</p>
### tuples

The main difference between Lists and tuples is the fact that tuples is an immutable type of data, making it faster to use.

<div class = "python">
```{python tup,echo=T,collapse = TRUE}
# Python

a = (1, 2, 3)
a.count(2) # count elements of the tuple which are exactly equal to 2

a
# access the element of a tuple
a[0]
a.index(3)
a[1:]
a[:1]
a[0:-1]
a[:]

```
```{python tup2,echo = T,collapse = TRUE}

# modify
a += (4,5) 
a
a = a*2 # replicate the tuple n times
len(a) # number of elements in the tuple

```
```{python tup3,echo = T,collapse = TRUE}
# immutable

a = (1,2,3)
b = a
b += (4,5)
a

```
</div>


<p>&nbsp;</p>
### Dictionaries

Dictionary refers to a way of storing data that is not sorted. It works with key and value associate with this key.


<div class = "python">
```{python dic,echo=T,collapse = TRUE}
# Python

a = {'a':1, 'b':2, 'c':3}
# access the element of a dictionary
a.keys()
a['a']
a.values()
a.items()
a.get('a')
a.get('d',4) # set to 4 if the key 'd' is not detected

a.pop('a') # pop will use the corresponding value to the key a and remove the pair (key, value).
a
a.popitem('b')
a
```
```{python dic2,echo = T,collapse = TRUE}
# modify
a['a'] =1 
a.setdefault('d',0) # create new item with a default value

b = {'d':4,'e':5}
a.update(b) # update values from other dict
a
a.clear() # remove all items

```
```{python dic3,echo = T,collapse = TRUE}
# mutable

a = {'a':1, 'b':2, 'c':3}
b = a
b['b'] = [12,14]
a
```

</div>


<p>&nbsp;</p>
### Sets

Sets are unordered collection of unique elements. If we give to a set multiple time the same element, it will automatically delete dupplicated values.


<div class = "python">
```{python set,echo=T,collapse = TRUE}
# Python

a = {1, 2, 3}
a
# access the element of a set

a[0] # since it unordered, we can not access to a given element of a set


```
```{python set2,echo = T,collapse = TRUE}
# modify
b = {3,4,5}
a.update(b)  # update values from other set
a

```
```{python set3,echo = T,collapse = TRUE}
# mutable

a = {1, 2, 3}
b = a
b.update([12,14])
a
```

</div>


<p>&nbsp;</p>
### Arrays

In order to manipulate arrays in Python we need to use package numpy, this package is very usefull and will be covered in other chapters.

First see that we can make an array using R by starting from a vector.
R represents all arrays in column-major order, which is not the case in python.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ar,echo=T,collapse = TRUE}
# Python

import numpy
arr = numpy.array([[1,4],[2,5],[3,6]])
arr

type(arr)

vec = [1,2,3,4,5,6]

arr = numpy.reshape(vec,(3,2))
arr
arr = numpy.reshape(vec,(3,2), order = 'F')
arr
```


</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-ar2,echo = T,collapse = TRUE}
# R
arr = as.array(rbind(c(1,4),c(2,5),c(3,6)))
arr

vec = c(1,2,3,4,5,6)

arr = array(vec,dim = c(3,2))
arr
```
</div>
</div>
</div>

As you can see, R and Python does not store vector in the same way: R is storing them by column while python does it by row. Using "order = 'F'" alows us to store vector into matrix by column.


<div class = "python">
```{python ar2,echo=T,collapse = TRUE}
# Python

# access the element of an array

arr[0] # access directly to the raw 1


```
```{python ar3,echo = T,collapse = TRUE}
# modify
vec = [7,8]
arr = numpy.insert(arr, len(arr),vec,axis = 0)  # update values from other set
arr

```
```{python ar4,echo = T,collapse = TRUE}
# mutable

arr2 = arr
arr2[0] = [12,14]
arr
```

</div>


<p>&nbsp;</p>

### Data Frame

Pandas Data Frames are also very common data-type in Python. The package Pandas is also view deeper in following chapters.

<div class = "python">
```{python pd,echo = T,collapse = TRUE}
# Python

import pandas

df = pandas.DataFrame(arr)
df

vec = [1,2,3,4,5,6]

df = pandas.DataFrame({'vec':vec,'vec1':range(2,8)})
df
```

</div>


<div class = "python">
```{python,echo=T,collapse = TRUE}
# Python

# access the element of a Pandas Data Frame

df['vec'] 

```
```{python,echo = T,collapse = TRUE}
# modify
vec2 = range(3,9)
df['vec2'] = vec2 # add values from other vector
a

```
```{python,echo = T,collapse = TRUE}
# mutable

df2 = df
df['vec'][0] = 30
df2
```

</div>


<p>&nbsp;</p>

# Control flow

There are two main control flow tools: choices, and loops. 
'Choices' are very useful for establishing rules or conditions. 'Choices' can be used to modify a value according to a certain condition, generally it allows to launch certain actions in specific cases.
The 'loops' when it allows to execute sequentially actions, it can be to interactively modify an object, more generally it allows to launch a procedure several times. For example, we can create N similar objects but we can also modify the N lines of an object.


## Choices



<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

# if, elif, else

n = 12

if n%2 == 0 :  
  print('n is an even number')
  



if n != int(n): 
  print('n is not a integer')
elif  n%2 == 0 :
  print('n is an even number')
else:
  print('n is not an even number')
  
  

  
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R

# if, elif, else

n = 12L

if(n%%2 == 0){  
  print('n is an even number')
}
  



if(!is.integer(n)){ 
  print('n is not a integer')
  } else if(n%%2 == 0){
    print('n is an even number')
    } else {
      print('n is not an even number')
      }


  
```
</div>
</div>
</div>
 


<p>&nbsp;</p>
## Loops

Using 'Loops', we iterate over a predefined number of iterations. Sometimes we do not know how many steps we need to perform a given task. Imagine that we want to optimize a function, we don't know how many steps we need until reaching an optimum, but we can set a condition for which we will consider that the algorithm converged. For this kind of exercice we can use 'while' loops, it will iterate until a given condition is satisfied.


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

seq = [1,2,None,4,None,6]
total = 0

for val in seq:
  if val is not None:
    total += val
    
total
```

```{python,echo = T,collapse = TRUE}
# Python 
import random

total = 0
while total < 1:
  rnd = random.gauss(mu = 0, sigma = 1)
  if rnd < 0:
    pass
  else:
    total += rnd
  
total

```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
seq = c(1,2,NA,4,NA,6)
total = 0

for(val in seq){
  if(!is.na(val)){
    total = total + val
  }
}
    
total
```


```{r,echo = T,collapse = TRUE}
# R

total = 0

while(total < 1){
  rnd = rnorm(n=1, mean = 0, sd = 1)
  if(rnd < 0){
    # in R we can specify nothing instead of specifiying 'pass'
    } else {
    total = total + rnd }
}
  
total
```

</div>
</div>
</div>

<p>&nbsp;</p>
## List, Set and Dict comprehensions (Python)

List comprehension is very common and appreciate in the python language features, think of it as a loop for which we will directly store output in a list, set, or dict. we can use it as a filter for exemple.


<div class = "python">
<div class = "row">
<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# List
import time

lst = [1,2,3,4]

t = time.time() 
results = []
for val in lst:
  if val > 2:
    results.append(val)
time.time()-t

results

t = time.time() 
# this loop their will produce the same output than a using List comprehension.
results = [val for val in lst if val>2]
time.time()-t

results
```
</div>
<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# Set
import time

st = {1,2,3,4}

t = time.time() 
results = set([])
for val in st:
  if val > 2:
    results.add(val)
time.time()-t

results

t = time.time() 

# this loop their will produce the same output than a using Set comprehension.
results = {val for val in st if val>2}
time.time()-t

results
```

</div>

<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# Dict
import time

dct = {'a':1,'b':2,'c':3,'d':4}

t = time.time() 
results = dict([])
for val in dct:
  if dct[val] > 2:
    results.update({str(val): dct[val]})
time.time()-t

results

t = time.time() 
# this loop their will produce the same output than a using Dict comprehension.
results = {str(val): dct[val] for val in dct if dct[val]>2}
time.time()-t

results
```

</div>
</div>
</div>










<p>&nbsp;</p>
# Functions


Functions are very important in both languages, being able to code ourselves things are sometimes more efficient than looking for a package and understand it.  
Being able to write our own functions gives us more flexibility and a better understanding of what we are actually doing.
It is important to not reinvent the wheel, but it is important to be able to construct it ourself. Read documentation carefully while using packages, sometimes packages are misleading and we can spend a lot of time understanding how they work. Look at the packages source code when we are not sure of what the function does behind.  
The full power of programming comes is the fact that we can be autonomous by reading, modifying and writing codes and pre-existing codes.



<p>&nbsp;</p>
## General Functions 


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 
seq = [1,2,None,4,None,6]*120

def clean_sum(seq):
  total = 0

  for val in seq:
    if val is not None:
      total += val
  return total
  
t = time.time()
clean_sum(seq = seq)
time.time() - t


def clean_sum2(seq):
  total = sum(filter(None,seq))#[val for val in seq if val is not None])
  return total
  
t = time.time()
clean_sum2(seq = seq)
time.time() - t


```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
seq = rep(c(1,2,NA,4,NA,6),120)


clean_sum <- function(seq){
  total = 0
  for(val in seq){
    if(!is.na(val)){
      total = total + val
    }
  }
  return(total)
}

t = Sys.time()
clean_sum(seq = seq)
Sys.time() - t



clean_sum2 <- function(seq){
  total <- sum(na.rm(seq))
  return(total)
}

t <- Sys.time()
clean_sum(seq = seq)
Sys.time() - t



```


</div>
</div>
</div>


Functions in Python are objects, they can have attributes and methods like objects. The functions can have data variables and even functions written inside of them.  
Suppose we want to apply several transformation to data, we can create multiple functions to do the different tasks that we want to perform. We can stock functions in list a apply them sequentially very easily.  
Note that we can separate the output into several variables by specifiying the different values before the assignation. We cannot do that in R directly bt the zeallot package can tackle this problem.



<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

def add_two(nb):
   nb = [i+2 for i in nb]
   return nb
  
def square_nb(nb):
   nb = [i**2 for i in nb]
   return nb

def global_function(nb):
   for function in func_list:
      nb = function(nb)
   return nb

func_list = [add_two,square_nb]

x1, x2 = global_function([2,3])
x1
x2
```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
library(zeallot)

add_two <- function(nb){
  nb = nb + 2
  return(nb)
}
  
square_nb <- function(nb){
  nb = nb**2
  return(nb)
}

global_function <- function(nb){
  for(func in func_list){
    nb = func(nb)
      }
  return(nb)
}

func_list = c(add_two,square_nb)

c(x1, x2) %<-% global_function(c(2,3))
x1
x2
```


</div>
</div>
</div>


<p>&nbsp;</p>

## Error and Exception Handling

Most of the time, writing a function is very task-specific, but since data are often dirty and incomplete, our given task may not work. It is important to understand why a given function can produce an error. Being sure that this and this case can leads to errors is not so easy, but it's important not to be too flexible to keep. In those examples, we are going to see how setting the flexibility of the function can leads to different results that are more or less efficient.

Look before we leap vs Easier to ask for forgiveness than permission.

If we expect that 99 % of the time result will actually contain something iterable, we would use the try/except approach. It will be faster if exceptions really are exceptional. If result is None more than 50 % of the time, then using 'if' is probably better.
So, whereas an 'if' statement always costs us, it's nearly free to set up a try/except block. But when an Exception actually occurs, the cost is much higher.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 
import numpy as np

# Let's create a function that create a dirty list
def create_dirty_list(None_prop):
  list_ = list()
  for i in range(10000):
    if i<None_prop*10000:
      list_.append([None])
    else:
      values = random.sample(range(1, 1000), random.sample(range(1, 50),1)[0])
      # introduce some character 
      if i%10==0 :
        values = [str(i) for i in values]
      list_.append(values)
  return list_

# randomly take two values to compute the ratio 
def calc1(values):
  output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
  return output

list_ = create_dirty_list(0.5)
# Store it in a list
results = [calc1(values) for  values in list_]

```

```{python,echo = T,collapse = TRUE}
# Let's change the function

def calc2(values):
  if not any(value is None for value in values): 
    output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
    return output

results = [calc2(values) for values in list_]

```

```{python,echo = T,collapse = TRUE}
# Let's change the function

def calc3(values):
  if not any(value is None for value in values): 
    if all(isinstance(value,int) for value in values):
      output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
      return output 

results = [calc3(values) for values in list_]

```

```{python,echo = T,collapse = TRUE}
# using try

def calc_try(values):
  try:
    output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
  except:
    output = None
  return output
    
results = [calc_try(values) for values in list_]
```

 - Compare execution time 

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.01)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.33)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.75)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

</div>
</div>

<div class = "col-md-6">
<div class = "r">

```{r,warning=F}
# R
# Let's create a function that create a list 
create_dirty_list <- function(None_prop){
  list_ <- list()
  for(i in seq(1,10000,1)){
    if(i<None_prop*10000){
      list_[[i]] <- c(NA)
      } else {
      values <- sample(seq(1,10000,1),round(runif(1,1,50)))
      # introduce some character 
      if(i%%10==0){
        values <- as.character(values)
      list_[[i]] <- values
      }
      }
  }
  return(list_)
  }

# randomly take two values to compute the ratio 
calc1 <- function(values){
  output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
  return(output)
}

list_ <- create_dirty_list(0.5)
# Store it in a list
results <- lapply(list_,FUN = calc1)

```

```{r,warning=F}
# Let's change the function

calc2 <- function(values){
  if(!any(is.na(values))){
    output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
    return(output)
  }
}


results <- lapply(list_,FUN = calc2)
```

```{r,warning=F}
# Let's change the function

calc3 <- function(values){
  if(!any(is.na(values))){
    if(is.numeric(values)){
      output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
      return(output)
    }
  }
}



results <- lapply(list_,FUN = calc3)

```

```{r,echo = T,warning=F}
# using try

calc_try <- function(values){
  output <- tryCatch(
    {
      values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
    }
    ,error=function(e){
      return(NA)
    })
  return(output)
}
    
results <- lapply(list_,FUN = calc_try)
```

 - Compare execution time

```{r,echo = T,warning=F}
 
list_ <- create_dirty_list(0.01)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```

```{r,echo = T,warning=F}
# compare it 
list_ = create_dirty_list(0.33)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```

```{r,echo = T,warning=F}
# compare it 
list_ = create_dirty_list(0.75)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```
</div>
</div>
</div>



<p>&nbsp;</p>

# Exercises

## Exercise 1


If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.


## Exercise 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

## Exercise 3



You are given the following information, but you may prefer to do some research for yourself.

- 1 Jan 1900 was a Monday.
- Thirty days has September,
    April, June and November.
    All the rest have thirty-one,
    Saving February alone,
    Which has twenty-eight, rain or shine.
    And on leap years, twenty-nine.
- A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

How many Sundays fell on the first of the month during the twentieth century (1 Jan 1901 to 31 Dec 2000)?



<p>&nbsp;</p>

<a href="">Possible solutions in R and Python without vectorisation</a>

