---
title: "Basics"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
library(reticulate)

#  md_document:
#    variant: markdown_github
#
#
#<div class = "row">
#<div class = "col-md-6">
#<div class = "python">
#  
#
#</div>
#</div>
#<div class = "col-md-6">
#<div class = "r">  
#  
#  
#</div>
#</div>
#</div>


#<style>
#div.blue pre { background-color:lightblue; }
#div.blue pre.r { background-color:blue; }
#</style>
#
```

<style>
div.python pre { background-color: #e9f7f5
; }
</style>

<style>
div.r pre { background-color: #fff5fd; }
</style>


<p>&nbsp;</p>

This chapter introduces the basics of R and Python, focusing on the different types of objects that can be used in each language, as well as control flow and function writing. The chapter concludes with three exercises from the Euler Project. Students are encouraged to work in pairs and solve at least one exercise in their chosen language, comparing different approaches.



<p>&nbsp;</p>



# R/Python

Data science languages such as R and Python are highly effective for business analysis and machine learning. Both languages offer a wide range of tools and libraries for data wrangling, preprocessing, and machine learning, making them ideal for reproducible research. Additionally, the ability to combine languages can open up new possibilities for even more advanced analysis. Despite their similarities, each language also has its own unique features and strengths, making it worth considering both options when choosing a language for a specific project.


## R Strengths

The R programming language has its origins in the S language, which was developed by statistician John Chambers at Bell Labs in 1976. The R language was later developed at the University of Auckland, New Zealand, to expand upon the capabilities of S. Both S and R were created by researchers for the purpose of more efficiently conducting research and communicating results.

R is a powerful language for statistics, data analysis, data exploration, and data visualization. It is equipped with tools for reporting and communication, such as RMarkdown and Shiny.

In recent years, R has seen significant growth with the emergence of the tidyverse (tidyverse.org), a collection of tools with a common programming interface that use functional verbs to perform intuitive operations, connected by the pipe operator. The tidyverse is particularly advantageous as it makes data mining more efficient and allows for easy iteration through exploratory analysis. The interface is designed in a way that it makes it easy to write a paragraph describing the steps that you intend to take with the data.

## Python Strengths

Python is a general-purpose programming language that was created by computer scientist Guido van Rossum in 1991. The language was designed to be highly readable and to encompass a wide range of programming paradigms. One of Python's key strengths is its flexibility, which allows it to handle a variety of tasks such as web frameworks, database connectivity, networking, web scraping, text and image processing, and many other features that are useful in machine learning.

Python is based on computer science and mathematics and boasts one of the largest ecosystems of any programming language, with over 100,000 open-source libraries. This makes it an ideal choice for those who value flexibility.

Python also has a rich set of data science libraries, including Scikit Learn, the most popular machine learning library, which is easy to learn, supports pipelines to simplify the machine learning workflow and has most of the algorithms you need in one place. Another common library is TensorFlow, developed by software engineers at Google for deep learning and commonly used for image recognition and natural language processing tasks. PyTorch, developed by Facebook is also a popular deep learning framework and a concurrent of Tensorflow and Keras (which is designed for efficiently building neural networks).
 
- Exploration is done using R for its readability and efficiency in the tidyverse and datatable.

- Machine learning is done using Python for its capability in the Scikit Learn, TensorFlow, Keras and PyTorch pipeline.

- R is used for communication because of the advanced reporting utilities such as RMarkdown, Shiny, and ggplot2 for visualization.

# Operations

## Arithmetic Operations
Python and R work as classical calculator, using "+", "-", "*" and "/" we can do arithmetic operations in both languages.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ope}
# Python 

1+2
1-2
1/2
1*2

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-ope, collapse = TRUE}
# R

1+2
1-2
1/2
1*2
```

</div>
</div>
</div>

<p>&nbsp;</p>

We can also apply exponentiation, Modulo and floor division easily in both language. Note that in R we can write the exponentiation "^". It is used for exclusive-or (XOR) in Python.


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ope2,echo = T,collapse = TRUE}
# Python 

2**8 # exponentiation
2^8 == 2**8 # False
8%3 # modulo
8//3 # floor division
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-ope2,echo=T,collapse = TRUE}
# R

2**8 # exponentiation
2^8 == 2**8 # TRUE
8%%3 # modulo
8%/%3 # floor division

```

</div>
</div>
</div>
Operator (R) | Operator (Python) |	Description
-|:-:|-:
+ | + |	Addition
– | – |	Subtraction
* | * |	Multiplication
/ | / |	Division
^ / ** | ** |	Exponent
%% | % |	Modulo
%/% | // |	floor Division


<p>&nbsp;</p>
##  Comparison Operators

To compare values, we use comparison operators to determine if a value is equal to, not equal to, greater than, etc.

These operators are the same in both languages.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python comp,echo = T,collapse = TRUE}
# Python 

2==8
2!=8
2<8
2>8
2<=8

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-cromp,echo=T,collapse = TRUE}
# R

2==8
2!=8
2<8
2>8
2<=8
```
</div>
</div>
</div>

Operator (R/Python) |	Description
-|-:
\< 	| Less than
\> 	| Greater than
\<= 	| Less than or equal to
\>= 	| Greater than or equal to
== 	| Equal to
!= 	| Not equal to

<p>&nbsp;</p>

## Logical operators


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python logope,echo = T,collapse = TRUE}
# Python 
x = [True,True]
y = [True,False]

not x[0]
x and y
x or y
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-logope,echo=T,collapse = TRUE}
# R

x = c(TRUE,TRUE)
y = c(TRUE,FALSE)

!x[1]
x & y
x && y
x | y
x || y
```
</div>
</div>
</div>

Operator (R)| Operator (Python)|	Description
-|:-:|-:
! | not |	Logical NOT
& | and |	Element-wise logical AND
&& |  |Logical AND
\| |or|	Element-wise logical OR
\|\|  | |	Logical OR


<p>&nbsp;</p>
## Membership operators

Python has 'in' and 'not in' operators for testing membership in a sequence, while R uses '%in%'. However, it's important to note that these operators may not work the same way in both languages, as in Python a list of characters can be directly decomposed into a character list which is not the case in R. For example, in Python we can use the 'in' operator to check if 'Hello' is in 'Hello world', whereas in R, the '%in%' operator checks for an exact match between a string and an element.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python relope,echo = T,collapse = TRUE}
# Python 
x = 'Hello world'
y = {1:'a',2:'b'}

print('world' in x)

print(1 in y)

print('a' in y)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-relope,echo=T,collapse = TRUE}
# R

x = 'Hello World'

print('Hello' %in% x)
stringr::str_detect(x,'Hello')
```
</div>
</div>
</div>



<p>&nbsp;</p>

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python relope2,echo = T,collapse = TRUE}
# Python 
x = ['Hello','World']

print('Hello' in x)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-relope2,echo=T,collapse = TRUE}
# R

x = c('Hello','World')

print('Hello' %in% x)
stringr::str_detect('Hello',x)
```
</div>
</div>
</div>


<p>&nbsp;</p>

# Objects & Variables

Both Python and R do not require declaring a variable before assigning a value to it. Variables can be thought of as names that refer to an object. However, there is a difference in the way objects and variables are stored in the computer's memory in Python and R.


<p>&nbsp;</p>

## Some differences between R and Python

While Python is widely appreciated for being a versatile language with an easy-to-understand syntax, R's functionalities are developed by statisticians, giving it specific advantages in certain fields.

In R, a variable and an object are the same thing, they refer to the same entities. If we assign a new variable to a variable that already exists, it will refer to two different objects. In Python, things are more conventional, but it's important to note that, even though variables refer to an object, they are still two separate entities. This means that two variables can refer to a single object. If two variables point to the same object, any modifications made using one variable will also be reflected when using the other variable.

We should generally avoid using for loops in R as they are very slow because they execute a function call with every iteration. Instead, we should use vectorization and the apply family of functions. Vectorization is crucial for fast code in R. Assuming we vectorize both our R and Python codes (and other factors are equal), we should expect similar performance in terms of speed. However, for data that exceeds the limit of memory, R may not be the best choice.



<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python ov,echo = T,collapse = TRUE}
# Python 
import numpy as np
only_nb = [1,2,3,4]
np.std(only_nb)
np.std(only_nb,ddof=1)
with_none = [1,2,3,None]
np.std(with_none)
with_char = [1,2,'3',4]
np.std(with_char)

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-ov,echo=T,collapse = TRUE}
# R
only_nb <- c(1,2,3,4)
sd(only_nb)
with_na <- c(1,2,3,NA)
sd(with_na)
with_char <- c(1,2,'3',4)
sd(with_char)

with_char <- c(1,2,'a',4)
sd(with_char)

```
</div>
</div>

<p>&nbsp;</p>

- When computing a standard deviation in R, it does not show an error for missing values (NA) and returns NA. It also attempts to convert strings to numeric values if possible. If this is not possible, R will return a NA. This is not an error message, but a warning message. It's important to be careful when writing code as R can continue calculations and return NA without stopping the process, unlike Python.

- R calculates the standard deviation with N-1 as the denominator, while numpy uses N. To obtain the same results when using R, we need to use 'ddof=1' as an argument in the std function.

<p>&nbsp;</p>
## Textual and numerical variables

In R, textual data is referred to as 'character', while in Python it is called a 'string', abbreviated as 'str'. Both languages use " or ' when defining a textual variable, but it's also possible to set a variable as a textual variable if needed.

Numerical variables are divided into three types. In Python, we have integer, float, and complex, while in R, the equivalent of a float is called 'double', referring to double-precision floating point. Without going into detail, this means that the number has a precision of up to 15 decimal places. Float can refer to a precision of 7 decimal places, but in Python, the data-type 'float' has a precision of 15 digits. There are ways to achieve higher precision using Numpy for example.

Finally, it's easy to check the data-type of a variable by using the command 'type' in Python or 'typeof' in R.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python tnv,echo = T,collapse = TRUE}
# Python 

import sys 
sys.float_info.dig # number of decimals

a = 1
type(a)

b = 1.1
type(b)

c = 1.1+2j
type(c)

d = 'd'
type(d)

# change the type
e = 2
f = str(e)
f
float(f)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r r-tnv,echo=T,collapse = TRUE}
# R

a = 1
typeof(a)
a_int = 1L
typeof(a_int)

b = 1.1
typeof(b)

c = 1.1+2i
typeof(c)

d = 'd'
typeof(d)

# change the type
e = 2
f = as.character(e)
f
as.numeric(f)
```
</div>
</div>
</div>

<p>&nbsp;</p>

## Other Data-Type (R)

There are many data types in both languages that will be covered throughout this course, for now, we will focus only on basic data types that are available without using any additional packages or libraries. In R, common data types include Vector, Matrix, List, and Dataframe.

<p>&nbsp;</p>

### Vectors and Matrices

<div class = "r">

<div class = "row">
<div class = "col-md-6">

```{r r-vm}
# R

vec = c(1, 2, 3)
length(which(vec==2)) # count elements of the list which are exactly equal to 2

vec = sort(vec,
           decreasing = TRUE)
vec
# access the element of a list
vec[1]
vec[which(vec==3)]
vec[2:length(vec)]
vec[length(vec):2]
vec[-1] # throwaway value


```

</div>
<div class = "col-md-6">

```{r r-vm2}
# R

mat = matrix(c(1, 2, 3, 4, 5, 6),3,2)
dim(mat)
mat

length(which(mat==2)) # count elements of the list which are exactly equal to 2

vec = sort(mat,
           decreasing = TRUE)
vec
# access the element of a list
mat[1] # first element
which(mat==3,
      arr.ind = TRUE) # get row and columns for each value that match

mat[2:length(mat)] # gives us a vector
mat[2:dim(mat)[1],]
mat[length(mat):1]

mat[1,-1] # throwaway values


```

</div>
</div>

```{r r-vm3}
# modify

b = rep(0,3)
vec = c(vec,b)
vec
mat_2 = rbind(vec,vec)
mat_2 
mat_2 = cbind(mat_2,mat_2)
mat_2

t_mat = t(mat) # transposition
t_mat[2,3] = as.character(t_mat[2,3]) # all values turn to character


i_mat = diag(1, 3)
i_mat


```


</div>

<p>&nbsp;</p>

### Data frame

A Data frame in R is a matrix that can store different types of data such as characters, numeric or factors. Unlike a matrix, it does not have a fixed dimension, which means that we can easily create a new column using the operator '$'.

<div class = "r">

```{r r-df,echo=T,collapse = TRUE}
# R

df = data.frame(chr = letters[1:5], # letters is a vector of letter provided directly in R
                num = seq(1,5,1),
                fac = factor(c(rep('a',3),rep('b',2)))) 
class(df$chr) # automatically converted to factor, which can be very inefficient ! UP: Not any more with R >= 4.0 !
df$chr = as.character(df$chr)
class(df$chr)
colnames(df)
df
# access the element of a variable in a df
df$chr[3]
which(df==3,
      arr.ind = TRUE) 

df[1:2] # gives us a df with the first two columns !
df[3:dim(df)[1],]

df[1,-1] # throwaway values

```
```{r r-df2,echo = T,collapse = TRUE}
# modify

b = rep(0,3)
df = rbind(df,b) # introduce a new row in the data frame, note that we get a NA in the factor column because '0' is not a level
df$new_col = rep(2,6) # we can introduce new columns by specifying the name and set a value


df
t_df = t(df) # transposition convert the df into a matrix
class(t_df)
```
</div>

<p>&nbsp;</p>

### Arrays

Arrays is a generalization of the matrix, meaning that we can have a matrix with more than two dimensions. We can modify them as we modify a matrix.

<div class = "r">

```{r r-ar,echo=T,collapse = TRUE}
# R

arr = array(seq(1,12,1), dim = c(2,2,3)) # create a matrix with 3 dimension
arr
# access the element of a dictionary
arr[1,,]
arr[,1,]
arr[,,1]
```
</div>


<p>&nbsp;</p>

### Lists 

A list permits us to assemble an assortment of disconnected objects behind one object, it is also possible to store objects which can be of different types.


<div class = "r">

```{r r-lst,echo=T,collapse = TRUE}
# R

lst = list(df = df,
           mat = mat,
           arr = arr) # create a matrix with 3 dimension

lst
lst[[1]] # first element of the list, here it is the df
lst[[1]][1,] # first row of first element of the list
```
</div>



<p>&nbsp;</p>

## Other Data-Type (Python)


In Python, the most common data types are List, Tuple, Sets, and Dictionary. It's important to note that when modifying a variable, we always use '='. Keep in mind that using the '.' in Python can modify the object behind the variable.


<p>&nbsp;</p>

### Lists 

Keep in mind that Lists are mutable, as discussed in the Section "Difference between R and Python"

<div class = "python">
```{python lst,echo = T,collapse = TRUE}
# Python

a = [1, 2, 3]
a.count(2) # count elements of the list which are exactly equal to 2

a.sort(reverse = True)
a
# access the element of a list
a[0]
a.index(3)
a[1:]
a[:1]
a[0:-1]
a[:]

```
```{python lst2,echo = T,collapse = TRUE}
# modify

b = [0, 0, 0]
list(zip(a,b)) # zip will pairs the ellements, it works also with more than 2 element ex: zip(a,b,c)
a.append(b)
a
a[4:5] = ['a','b']
a
a.extend([4, 5, 6])
a
a += [7,8] # works as extend
a
a.insert(2,[1,2])
a
a.remove(b)
a

a = a*2 # replicate the list n times
len(a) # number of elements in the list

```

```{python lst3,echo = T,collapse = TRUE}
# mutable

a = [1,2,3]
b = a
b[0] = 12
a

```
</div>



<p>&nbsp;</p>
### tuples

The main difference between Lists and tuples is the fact that tuples is an immutable type of data, making it faster to use.

<div class = "python">
```{python tup,echo=T,collapse = TRUE}
# Python

a = (1, 2, 3)
a.count(2) # count elements of the tuple which are exactly equal to 2

a
# access the element of a tuple
a[0]
a.index(3)
a[1:]
a[:1]
a[0:-1]
a[:]

```
```{python tup2,echo = T,collapse = TRUE}

# modify
a += (4,5) 
a
a = a*2 # replicate the tuple n times
len(a) # number of elements in the tuple

```
```{python tup3,echo = T,collapse = TRUE}
# immutable

a = (1,2,3)
b = a
b += (4,5)
a

```
</div>


<p>&nbsp;</p>
### Dictionaries

Dictionary refers to a way of storing data that is not sorted. It works with key and value associate with this key.


<div class = "python">
```{python dic,echo=T,collapse = TRUE}
# Python

a = {'a':1, 'b':2, 'c':3}
# access the element of a dictionary
a.keys()
a['a']
a.values()
a.items()
a.get('a')
a.get('d',4) # set to 4 if the key 'd' is not detected

a.pop('a') # pop will use the corresponding value to the key a and remove the pair (key, value).
a
a.popitem('b')
a
```
```{python dic2,echo = T,collapse = TRUE}
# modify
a['a'] =1 
a.setdefault('d',0) # create new item with a default value

b = {'d':4,'e':5}
a.update(b) # update values from other dict
a
a.clear() # remove all items

```
```{python dic3,echo = T,collapse = TRUE}
# mutable

a = {'a':1, 'b':2, 'c':3}
b = a
b['b'] = [12,14]
a
```

</div>


<p>&nbsp;</p>
### Sets

Sets are unordered collection of unique elements. If we give to a set multiple time the same element, it will automatically delete duplicated values.


<div class = "python">
```{python set,echo=T,collapse = TRUE}
# Python

a = {1, 2, 3}
a
# access the element of a set

a[0] # since it unordered, we can not access to a given element of a set


```
```{python set2,echo = T,collapse = TRUE}
# modify
b = {3,4,5}
a.update(b)  # update values from other set
a

```
```{python set3,echo = T,collapse = TRUE}
# mutable

a = {1, 2, 3}
b = a
b.update([12,14])
a
```

</div>


<p>&nbsp;</p>
### Arrays

To manipulate arrays in Python, we use the numpy package. This package is very useful and will be discussed in more detail in later chapters.

It's worth noting that in R, arrays are created using vectors and are stored in column-major order, which is different from how arrays are handled in Python.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python ar,echo=T,collapse = TRUE}
# Python

import numpy
arr = numpy.array([[1,4],[2,5],[3,6]])
arr

type(arr)

vec = [1,2,3,4,5,6]

arr = numpy.reshape(vec,(3,2))
arr
arr = numpy.reshape(vec,(3,2), order = 'F')
arr
```


</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r r-ar2,echo = T,collapse = TRUE}
# R
arr = as.array(rbind(c(1,4),c(2,5),c(3,6)))
arr

vec = c(1,2,3,4,5,6)

arr = array(vec,dim = c(3,2))
arr
```
</div>
</div>
</div>

As you can see, R and Python handle vector storage differently: R stores vectors by column while Python stores them by row. However, in Python, using the "order = 'F'" parameter allows vectors to be stored in matrices by column.


<div class = "python">
```{python ar2,echo=T,collapse = TRUE}
# Python

# access the element of an array

arr[0] # access directly to the raw 1


```
```{python ar3,echo = T,collapse = TRUE}
# modify
vec = [7,8]
arr = numpy.insert(arr, len(arr),vec,axis = 0)  # update values from other set
arr

```
```{python ar4,echo = T,collapse = TRUE}
# mutable

arr2 = arr
arr2[0] = [12,14]
arr
```

</div>


<p>&nbsp;</p>

### Data Frame

Pandas Data Frames are also very common data-type in Python. The package Pandas is also view deeper in following chapters.

<div class = "python">
```{python pd,echo = T,collapse = TRUE}
# Python

import pandas

df = pandas.DataFrame(arr)
df

vec = [1,2,3,4,5,6]

df = pandas.DataFrame({'vec':vec,'vec1':range(2,8)})
df
```

</div>


<div class = "python">
```{python,echo=T,collapse = TRUE}
# Python

# access element of a Pandas Data Frame

df['vec'] 

```
```{python,echo = T,collapse = TRUE}
# modify
vec2 = range(3,9)
df['vec2'] = vec2 # add values from other vector
a

```
```{python,echo = T,collapse = TRUE}
# mutable

df2 = df
df['vec'][0] = 30
df2
```

</div>


<p>&nbsp;</p>

# Control flow

In programming, there are two main control flow tools: conditional statements and loops.

Conditional statements, also known as choices, are useful for establishing rules or conditions. They allow for modifying a value according to a certain condition, and generally allow for certain actions to be taken in specific cases.

Loops, on the other hand, allow for sequential execution of actions. They can be used to interactively modify an object, and generally allow for a procedure to be executed multiple times. For example, we can use loops to create multiple similar objects, or to modify multiple lines in a single object.


## Choices



<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

# if, elif, else

n = 12

if n%2 == 0 :  
  print('n is an even number')
  



if n != int(n): 
  print('n is not a integer')
elif  n%2 == 0 :
  print('n is an even number')
else:
  print('n is not an even number')
  
  

  
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R

# if, elif, else

n = 12L

if(n%%2 == 0){  
  print('n is an even number')
}
  



if(!is.integer(n)){ 
  print('n is not a integer')
  } else if(n%%2 == 0){
    print('n is an even number')
    } else {
      print('n is not an even number')
      }


  
```
</div>
</div>
</div>
 


<p>&nbsp;</p>
## Loops

With 'loops', we iterate over a predefined number of iterations. However, in certain situations we may not know in advance how many iterations are required to complete a task.

For example, when trying to optimize a function, we may not know how many steps are needed to reach an optimum, but we can set a condition for when the algorithm is considered to have converged. In such cases, we can use a 'while' loop, which will iterate until a given condition is met.


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

seq = [1,2,None,4,None,6]
total = 0

for val in seq:
  if val is not None:
    total += val
    
total
```

```{python,echo = T,collapse = TRUE}
# Python 
import random

total = 0
while total < 1:
  rnd = random.gauss(mu = 0, sigma = 1)
  if rnd < 0:
    pass
  else:
    total += rnd
  
total

```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
seq = c(1,2,NA,4,NA,6)
total = 0

for(val in seq){
  if(!is.na(val)){
    total = total + val
  }
}
    
total
```


```{r,echo = T,collapse = TRUE}
# R

total = 0

while(total < 1){
  rnd = rnorm(n=1, mean = 0, sd = 1)
  if(rnd < 0){
    # in R we can specify nothing instead of specifying 'pass'
    } else {
    total = total + rnd }
}
  
total
```

</div>
</div>
</div>

<p>&nbsp;</p>
## List, Set and Dict comprehensions (Python)

List comprehension is very common and appreciate in the python language features, think of it as a loop for which we will directly store output in a list, set, or dict. we can use it as a filter for example.


<div class = "python">
<div class = "row">
<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# List
import time

lst = [1,2,3,4]

t = time.time() 
results = []
for val in lst:
  if val > 2:
    results.append(val)
time.time()-t

results

t = time.time() 
# this loop their will produce the same output than a using List comprehension.
results = [val for val in lst if val>2]
time.time()-t

results
```
</div>
<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# Set
import time

st = {1,2,3,4}

t = time.time() 
results = set([])
for val in st:
  if val > 2:
    results.add(val)
time.time()-t

results

t = time.time() 

# this loop their will produce the same output than a using Set comprehension.
results = {val for val in st if val>2}
time.time()-t

results
```

</div>

<div class = "col-md-6">
```{python,echo=T,collapse = TRUE}
# Python 
# Dict
import time

dct = {'a':1,'b':2,'c':3,'d':4}

t = time.time() 
results = dict([])
for val in dct:
  if dct[val] > 2:
    results.update({str(val): dct[val]})
time.time()-t

results

t = time.time() 
# this loop their will produce the same output than a using Dict comprehension.
results = {str(val): dct[val] for val in dct if dct[val]>2}
time.time()-t

results
```

</div>
</div>
</div>










<p>&nbsp;</p>
# Functions

Functions are an important aspect of both R and Python. Being able to write our own functions can be more efficient than searching for and understanding a pre-existing package.

Writing our own functions gives us more flexibility and a better understanding of what we are doing. However, it is important to not reinvent the wheel and instead use pre-existing packages when appropriate. It is crucial to carefully read the documentation when using packages, as they can be misleading and lead to a lot of time spent trying to understand how they work. It is also beneficial to look at the package's source code when unsure of what a function does behind the scenes.

The full power of programming comes from the ability to be autonomous by reading, modifying, and writing code, as well as utilizing pre-existing code.



<p>&nbsp;</p>
## General Functions 


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 
seq = [1,2,None,4,None,6]*120

def clean_sum(seq):
  total = 0

  for val in seq:
    if val is not None:
      total += val
  return total
  
t = time.time()
clean_sum(seq = seq)
time.time() - t


def clean_sum2(seq):
  total = sum(filter(None,seq))#[val for val in seq if val is not None])
  return total
  
t = time.time()
clean_sum2(seq = seq)
time.time() - t


```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
seq = rep(c(1,2,NA,4,NA,6),120)


clean_sum <- function(seq){
  total = 0
  for(val in seq){
    if(!is.na(val)){
      total = total + val
    }
  }
  return(total)
}

t = Sys.time()
clean_sum(seq = seq)
Sys.time() - t



clean_sum2 <- function(seq){
  total <- sum(na.rm(seq))
  return(total)
}

t <- Sys.time()
clean_sum(seq = seq)
Sys.time() - t



```


</div>
</div>
</div>

Functions in Python are objects and can have attributes and methods like any other object. Functions can also contain data variables and even other functions.

For example, if we want to apply multiple transformations to data, we can create separate functions for each task. These functions can be stored in a list and applied sequentially with ease.

It's worth noting that in Python, it is possible to unpack the output of a function into multiple variables by specifying them before the assignment. In R, this is not possible directly, but the zeallot package can be used to achieve this functionality.


<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 

def add_two(nb):
   nb = [i+2 for i in nb]
   return nb
  
def square_nb(nb):
   nb = [i**2 for i in nb]
   return nb

def global_function(nb):
   for function in func_list:
      nb = function(nb)
   return nb

func_list = [add_two,square_nb]

x1, x2 = global_function([2,3])
x1
x2
```

</div>
</div>

<div class = "col-md-6">
<div class = "r">
```{r,echo=T,collapse = TRUE}
# R
library(zeallot)

add_two <- function(nb){
  nb = nb + 2
  return(nb)
}
  
square_nb <- function(nb){
  nb = nb**2
  return(nb)
}

global_function <- function(nb){
  for(func in func_list){
    nb = func(nb)
      }
  return(nb)
}

func_list = c(add_two,square_nb)

c(x1, x2) %<-% global_function(c(2,3))
x1
x2
```


</div>
</div>
</div>


<p>&nbsp;</p>

## Error and Exception Handling

It is important to understand why a given function may produce an error. Determining the situations where errors may occur is not always easy, but it's important to not be too lenient in order to avoid errors. In the examples, we will see how the flexibility of a function can lead to different results, some of which may be more efficient than others.

For example, if we expect that 99% of the time the result will contain something iterable, we would use the try/except approach. This will be faster if exceptions are truly exceptional. However, if the result is None more than 50% of the time, then using an 'if' statement is probably a better approach.

While an 'if' statement always has a cost, setting up a try/except block is relatively inexpensive. However, when an exception does occur, the cost is much higher.

<div class = "row">
<div class = "col-md-6">
<div class = "python">
```{python,echo = T,collapse = TRUE}
# Python 
import numpy as np

# Let's create a function that create a dirty list
def create_dirty_list(None_prop):
  list_ = list()
  for i in range(10000):
    if i<None_prop*10000:
      list_.append([None])
    else:
      values = random.sample(range(1, 1000), random.sample(range(1, 50),1)[0])
      # introduce some character 
      if i%10==0 :
        values = [str(i) for i in values]
      list_.append(values)
  return list_

# randomly take two values to compute the ratio 
def calc1(values):
  output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
  return output

list_ = create_dirty_list(0.5)
# Store it in a list
results = [calc1(values) for  values in list_]

```

```{python,echo = T,collapse = TRUE}
# Let's change the function

def calc2(values):
  if not any(value is None for value in values): 
    output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
    return output

results = [calc2(values) for values in list_]

```

```{python,echo = T,collapse = TRUE}
# Let's change the function

def calc3(values):
  if not any(value is None for value in values): 
    if all(isinstance(value,int) for value in values):
      output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
      return output 

results = [calc3(values) for values in list_]

```

```{python,echo = T,collapse = TRUE}
# using try

def calc_try(values):
  try:
    output = values[random.sample(range(1, len(values)+1),1)[0]-1]/values[random.sample(range(1, len(values)+1),1)[0]-1]
  except:
    output = None
  return output
    
results = [calc_try(values) for values in list_]
```

 - Compare execution time 

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.01)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.33)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

```{python,echo = T,collapse = TRUE}
list_ = create_dirty_list(0.75)
t= time.time()
results = [calc3(values) for values in list_]
time.time()-t
t= time.time()
results = [calc_try(values) for values in list_]
time.time()-t

```

</div>
</div>

<div class = "col-md-6">
<div class = "r">

```{r,warning=F}
# R
# Let's create a function that create a list 
create_dirty_list <- function(None_prop){
  list_ <- list()
  for(i in seq(1,10000,1)){
    if(i<None_prop*10000){
      list_[[i]] <- c(NA)
      } else {
      values <- sample(seq(1,10000,1),round(runif(1,1,50)))
      # introduce some character 
      if(i%%10==0){
        values <- as.character(values)
      list_[[i]] <- values
      }
      }
  }
  return(list_)
  }

# randomly take two values to compute the ratio 
calc1 <- function(values){
  output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
  return(output)
}

list_ <- create_dirty_list(0.5)
# Store it in a list
results <- lapply(list_,FUN = calc1)

```

```{r,warning=F}
# Let's change the function

calc2 <- function(values){
  if(!any(is.na(values))){
    output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
    return(output)
  }
}


results <- lapply(list_,FUN = calc2)
```

```{r,warning=F}
# Let's change the function

calc3 <- function(values){
  if(!any(is.na(values))){
    if(is.numeric(values)){
      output <- values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
      return(output)
    }
  }
}



results <- lapply(list_,FUN = calc3)

```

```{r,echo = T,warning=F}
# using try

calc_try <- function(values){
  output <- tryCatch(
    {
      values[round(runif(1,1,length(values)))]/values[round(runif(1,1,length(values)))]
    }
    ,error=function(e){
      return(NA)
    })
  return(output)
}
    
results <- lapply(list_,FUN = calc_try)
```

 - Compare execution time

```{r,echo = T,warning=F}
 
list_ <- create_dirty_list(0.01)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```

```{r,echo = T,warning=F}
# compare it 
list_ = create_dirty_list(0.33)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```

```{r,echo = T,warning=F}
# compare it 
list_ = create_dirty_list(0.75)
t <- Sys.time()
results <- lapply(list_,FUN = calc3)
Sys.time()-t
t <- Sys.time()
results <- lapply(list_,FUN = calc_try)
Sys.time()-t

```
</div>
</div>
</div>



<p>&nbsp;</p>

# Exercises

## Exercise 1


If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.


## Exercise 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

## Exercise 3



You are given the following information, but you may prefer to do some research for yourself.

- 1 Jan 1900 was a Monday.
- Thirty days has September,
    April, June and November.
    All the rest have thirty-one,
    Saving February alone,
    Which has twenty-eight, rain or shine.
    And on leap years, twenty-nine.
- A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

How many Sundays fell on the first of the month during the twentieth century (1 Jan 1901 to 31 Dec 2000)?



<p>&nbsp;</p>

<a href="exo_chap_1.html">Possible solutions in R and Python without vectorisation</a>
