---
title: "Arrays"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
library(reticulate)

```

<style>
div.python pre { background-color: #d0f5a9; }
</style>

<style>
div.r pre { background-color: #dfdddf; }
</style>



This chapter deepens the notions of arrays and vectors seen in the 'Basics' chapter.
In python, numpy will be the main package to deal with arrays, while in R everything is contains in the base.
The main advantage of using vectors is to vectorize operation, instead of using loops to run operation on each elements of an object, we can simply perform it on blocks of data in a more efficient way.


<p>&nbsp;</p>
<p>&nbsp;</p>

let's first see an example to understand the benefit of using vectorized operation :

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

import numpy as np
import time 

arr = np.arange(2000000)

lst = list(range(2000000))

t = time.time()
arr_2 = arr*2
time.time() - t

t = time.time()
lst_2 = [i*2  for i in lst]
time.time() - t


```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

vec = c(1:2000000)

t = Sys.time()
vec_2 = vec*2
Sys.time() - t

t = Sys.time()
vec_2_loop <- rep(0,length(vec))
for(i in 1:length(vec)){
  vec_2_loop[i] = vec[i]*2
    }
Sys.time() - t

```

</div>
</div>
</div>


# Create arrays 

The NumPy library contains multidimensional array and matrix data structures, it provides methods to efficiently operate on it. If you do not have numpy on you computer, open a terminal and run ```pip install numpy```.

To use numpy one need to load the module by running ```import numpy as np```, you can just import numpy but then you will need to write numpy.something to perform calculation. In order to keep code standardized we simply say that we will refers to numpy using np.

Numpy allows to perform fast and efficient calculation. The main difference between Numpy and python list is that all elements in a numpy array must be homogenous, Numpy uses less memory to store data.

Functions | Tasks
-|-:
array	| Create numpy array
ndim |	Dimension of the array
shape	| Size of the array (Number of rows and Columns)
size	| Total number of elements in the array
dtype	| Type of elements in the array, i.e., int64, character
reshape	| Reshapes the array without changing the original shape
resize	| Reshapes the array. Also change the original shape
arange	| Create sequence of numbers in array
Itemsize	| Size in bytes of each item
diag	| Create a diagonal matrix
vstack	| Stacking vertically
hstack	| Stacking horizontally

- Starting from existing list or vector :

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

import numpy as np

arr = np.array([1, 2, 3, 4])
arr

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

arr = matrix(c(1, 2, 3, 4), 4, 1)
arr
```

</div>
</div>
</div>

- Create arrays with zero, one, sequence..


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

np.zeros([2, 3])


np.ones((2, 3))


np.arange(1, 7)


np.arange(1, 7).reshape(2, 3)

## remember from first chapter that array in R and Python aren't store the same way, set order = 'F' to get the same results


np.arange(1, 7).reshape([2, 3], order = 'F')

np.linspace(1, 4,num = 10)
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

matrix(rep(0,2*3), 2, 3)

matrix(rep(1,2*3), 2, 3)

seq(1,6)

matrix(seq(1,6),2,3)

seq(1, 4, length.out = 10)

```

</div>
</div>
</div>


## Random array 


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

#generate an array
arr_rd = np.random.randn(4,5)

arr_rd

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

#generate a matrix
mat_rd <- matrix(runif(4*5),4,5)

mat_rd


```

</div>
</div>
</div>

## Indexing and slicing

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

arr = np.arange(1,7).reshape(2,3)
arr[0]
arr[:2]
arr[1:]
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R


mat <- matrix(1:6,2,3,byrow = T)
mat[1,]
mat[1:2,]
mat[2:dim(mat)[1],]
```

</div>
</div>
</div>

## Shape and size
    


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

# .ndim gives the number of axes, or dimensions, of the array.
arr_rd.ndim
```

```{python}
# Python 

# .size gives the total number of elements of the array. 
arr_rd.size

```

```{python}
# Python 

# .shape display a tuple of integers with the number of elements stored along each dimension of the array
arr_rd.shape
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

# length gives the total number of elements of the array.
length(mat_rd)

```

```{r, collapse = TRUE}
#R

# dim display a vector of integers with the number of elements stored along each dimension of the array
dim(mat_rd)

```

</div>
</div>
</div>

# Modifiying arrays

## Add elements 

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 
arr

# /!\ using append it convert the array to a 1d array
arr_1d = np.append(arr, [7, 8, 9])
arr_1d

# using insert you can add rows and columns 
np.insert(arr, len(arr), [7, 8, 9], axis = 0)

np.insert(arr, 2, [7, 8], axis = 1)

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R
mat

rbind(mat,7:9)

cbind(mat,7:8)
```

</div>
</div>
</div>

- Combining/splitting arrays

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 
arr0 = np.zeros([2,3])
arr1 = np.ones([2,3])

arr01 = np.vstack([arr0,arr1])
arr01

#or 
np.concatenate([arr0,arr1], axis = 0)

np.hstack([arr0,arr1])

#or
np.concatenate([arr0,arr1], axis = 1)

np.hsplit(arr01,3)
np.vsplit(arr01,4)

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

mat0 = matrix(0,2,3)
mat1 = matrix(1,2,3)

mat01 = rbind(mat0,mat1)
mat01

cbind(mat0,mat1)

# horizontal
asplit(mat01,2)

#vertical
asplit(mat01,1)

```

</div>
</div>
</div>




## Delete elements


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

np.delete(arr,1 , axis = 1)

np.delete(arr,1 , axis = 0)

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

mat[,-2]

mat[-2,]
```

</div>
</div>
</div>

## Sorting


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

arr = np.random.randn(10)

arr
arr.sort()


```

```{python}
# Python 

arr = np.random.randn(4,3)

arr
arr.sort(1)


```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

arr <- runif(10)

arr
sort(arr)

```

```{r, collapse = TRUE}
# R

arr <- matrix(runif(4*3),4,3)

arr
apply(arr,MARGIN = 2, FUN = sort)

```

</div>
</div>
</div>

# Conditional Logic on array

Imagine two square matrices representing interaction between entities. Each matrix is a basic network: having interaction = 1, no interaction = 0.
You want to know if two entities that are linked in one network are also linked in an other network. Instead of checking each entities interaction sequentially in a loop, one can simply use matrix to do element-wise calculation and get the same answer.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

#generate a list
list_1 = [[0,1,0,1],[1,0,1,0],[0,1,0,1],[1,0,1,0]]
network_1 = np.array(list_1)

#generate a sequence
list_2 = [[0,1,0,0],[1,0,0,0],[0,0,0,1],[0,0,1,0]]
network_2 = np.array(list_2)

network_12 = network_1*network_2
network_12
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

#generate a sequence
list_1 <- list(c(0,1,0,1),c(1,0,1,0),c(0,1,0,1),c(1,0,1,0))

network_1 <- do.call(rbind,list_1)

#generate a sequence
list_2 = list(c(0,1,0,0),c(1,0,0,0),c(0,0,0,1),c(0,0,1,0))

network_2 <- do.call(rbind,list_2)

network_12 = network_1*network_2
network_12

```

</div>
</div>
</div>


## Select specific elements


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

network_12 > 0

network_12[network_12>0]
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R
 
network_12 > 0

network_12[network_12>0]
```

</div>
</div>
</div>


On thing that we often need is to modify specific value in a array that are satisfiying a condition. The first thing we want to know is which cells is satisfying the condition. In Python np.where is very common while in R we use the which function.

See how the execution time is changing depending on the size of the vector

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

arr = np.arange(200000)


t = time.time()
results = [(i if i%2==0 else 0) for i in arr]
time.time() - t


t = time.time()
results = np.where(arr%2!=0,0,arr)
time.time() - t
```

```{python}
# Python 

arr = np.arange(20000000)


t = time.time()
results = [(i if i%2==0 else 0) for i in arr]
time.time() - t


t = time.time()
results = np.where(arr%2!=0,0,arr)
time.time() - t
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

vec <- seq(1,200000)

t <- Sys.time()
results <- rep(0,length(vec))
for(i in 1:length(vec)){
  if(vec[i]%%2!=0){
    results[i] = 0
  } else {
    results[i] = vec[i]
  }
}
Sys.time() - t


t <- Sys.time()
results <- ifelse(vec%%2!=0,vec,0)
Sys.time() - t

```

```{r, collapse = TRUE}
# R

vec <- seq(1,20000000)

t <- Sys.time()
results <- rep(0,length(vec))
for(i in 1:length(vec)){
  if(vec[i]%%2!=0){
    results[i] = 0
  } else {
    results[i] = vec[i]
  }
}
Sys.time() - t


t <- Sys.time()
results <- ifelse(vec%%2!=0,vec,0)
Sys.time() - t


## other way using which

t <- Sys.time()
vec[which(vec%%2==0)] = 0
Sys.time() - t 

```

</div>
</div>
</div>

## Other conditional vectorization 

<div class = "row">
<div class = "r">

```{r,echo = T,collapse = TRUE}
# R
library(dplyr)
# Fast version of ifelse()
x <- c(-3:3, NA)
if_else(condition = x < 0,
        true      = "neg",
        false     = "pos",
        missing   = "NA")


# Vectorised ifelse statements


x <- 1:10
case_when(
  x %% 6 == 0 ~ "fizz buzz",
  x %% 2 == 0 ~ "fizz",
  x %% 3 == 0 ~ "buzz",
  TRUE ~ as.character(x)
)

```

</div>
</div>




# Algebra


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 
from numpy.linalg import inv, qr 
 
X = np.array([[0,1,5,1],[2,1,3,1],[2,1,9,6],[7,2,1,0],[8,3,5,5]]) 
Y = np.array([0,1,2,5,4])

X
X.T.dot(X) 
# same as 
XtX = np.dot(X.T,X)
# same as 
X.T @ X

# inverse Matrix
inv(XtX)

# Identity matrix
np.diag(np.ones(3))

```

```{python}
# Python 

XtY = np.dot(X.T,Y)

# OLS 
Beta = inv(XtX).dot(XtY)
Beta
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

 
X <- do.call(rbind,list(c(0,1,5,1),c(2,1,3,1),c(2,1,9,6),c(7,2,1,0),c(8,3,5,5))) 
Y <- c(0,1,2,5,4)
 
t(X)%*%X

XtX <- t(X)%*%X

# inverse Matrix
solve(XtX)

# Identity matrix
diag(3)
```

```{r, collapse = TRUE}
# R


XtY = t(X)%*%Y

# OLS 
Beta = solve(XtX) %*% XtY
Beta

```

</div>
</div>
</div>


# Operation on array

Operator	| Equivalent |Description
-|:-:|-:
+	| np.add	| Addition (e.g., 1 + 1 = 2)
-	| np.subtract	| Subtraction (e.g., 3 - 2 = 1)
-	| np.negative |	Unary negation (e.g., -2)
*	| np.multiply	| Multiplication (e.g., 2 * 3 = 6)
/	| np.divide	| Division (e.g., 3 / 2 = 1.5)
// |	np.floor_divide	| Floor division (e.g., 3 // 2 = 1)
**	| np.power	| Exponentiation (e.g., 2 ** 3 = 8)
%	| np.mod | Modulus/remainder (e.g., 9 % 4 = 1)


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

# nb of link
network_12.sum()/2
# same since there is only 0 and 1 
(network_12>0).sum()/2

# share of link 
network_12.mean()


# nb link by entities
network_12.sum(axis=1)

# share of link across entities (by columns)
network_12.mean(1)



```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

# nb of link
sum(network_12)
# same since there is only 0 and 1 
sum(network_12[which(network_12>0,arr.ind = T)]) # see the arr.ind to get the two coordinates

# share of link 
mean(network_12)


# nb link by entities
apply(network_12,MARGIN = 2,FUN = sum)

# share of link across entities (by columns)
apply(network_12,MARGIN = 2,FUN = mean)


```

</div>
</div>
</div>


## Apply familly (R)


The apply() family allows to manipulate slices of data from matrices, arrays, lists and dataframes in a repetitive way. It avoid explicit use of loop constructs. They act on an input list, matrix or array and apply a function with one or several optional arguments.

### apply

apply takes a matrix as input, transform it by row or by columns and returns a matrix

<div class = "r">

```{r, collapse = TRUE}
# R

mat_1 = matrix(1:(4*4),4,4)

# by row
apply(mat_1,MARGIN = 1,FUN = sum)

# by columns
apply(mat_1,MARGIN = 2,FUN = function(x){x**2})


```

</div>

### lapply


lapply takes a list as input, transform it and returns a list.


<div class = "r">

```{r, collapse = TRUE}
# R

list_1 = list(mat_1,seq(1,8,0.5))
list_1

# by row
lapply(list_1,FUN = function(x){x**2})

# by columns
lapply(list_1,FUN = sum)


```

</div>

### sapply

sapply takes a list as input, transform it and returns a matrix.

<div class = "r">

```{r, collapse = TRUE}
# R

list_1 = list(mat_1,seq(1,8,length.out = length(mat_1)))
list_1

# by row
sapply(list_1,FUN = function(x){x**2})

# by columns
sapply(list_1,FUN = sum)


```

</div>


### mapply

mapply is used for ‘multivariate’ apply. The main goal is to vectorize arguments to a function that is not usually accepting vectors as arguments. Depending on the size of the outputs it return a matrix or a list.

<div class = "r">

```{r, collapse = TRUE}
# R

# returns a matrix (all length.out = 5)
mapply(FUN = function(x,y,z){seq(x,y,length.out = z)},1,1:5,5)

# returns a list (length.out goes from 1 to 5)
mapply(FUN = function(x,y,z){seq(x,y,length.out = z)},1,1:5,1:5)


```

</div>

## Map (Python)

In python we can use ```map``` to do the same thing than apply familly in R with a bit of manipulation.

```map``` takes different kind of input, it transforms it by row if the object is an array and returns an iterator.You can then transform this into a list or an array.

<div class = "python">

```{python, collapse = TRUE}
# Python

arr_1 = np.arange(1,17).reshape([4,4], order = 'F')

# by row, by columns you need to tranpose the array
map(np.sum,arr_1)
list(map(np.sum,arr_1))
np.fromiter(map(sum,arr_1.T), dtype = np.int)

# with a list
lst_1 = [arr_1,np.linspace(1,8,num=arr_1.size)]
list(map(lambda x: x**2,lst_1))

```

</div>

- Pandas' apply function:

Using pandas one can also use apply functions, the difference with map is that by default it works by columns since variables in a df are store by colmuns.

<div class = "python">

```{python, collapse = TRUE}
# Python

import pandas as pd

arr_1 = np.arange(1,17).reshape([4,4], order = 'F')

#transform the array into a DataFrame
df_1 = pd.DataFrame(arr_1)
df_1


# Apply the function by column
df_1.apply(sum)


```

</div>


# Exercices

## Exercice 1

The series, $1^{1} + 2^{2} + 3^{3} + ... + 10^{10} = 10405071317$.

Find the last ten digits of the series, $1^{1} + 2^{2} + 3^{3} + ... + 1000^{1000}$.

## Exercice 2

Try to vectorize exercices from chapter 1, you can also compare it with apply/map functions. 

## Exercice 3

2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder.

What is the smallest positive number that is evenly divisible by all of the numbers from 1 to 20?

## Exercice 4
A palindromic number reads the same both ways. The largest palindrome made from the product of two 2-digit numbers is 9009 = 91 × 99.

Find the largest palindrome made from the product of two 3-digit numbers.

## Execrice 5

$2^{15} = 32768$ and the sum of its digits is 3 + 2 + 7 + 6 + 8 = 26.

What is the sum of the digits of the number $2^{1000}$?


<a href="exo_chap_2.html">Solutions</a>
