---
title: "Arrays/Vectors"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
library(reticulate)

```

<style>
div.python pre { background-color: #d0f5a9; }
</style>

<style>
div.r pre { background-color: #dfdddf; }
</style>



This chapter deepens the notions of arrays and vectors seen in the 'Basics' chapter.
In python, numpy will be the main package to deal with arrays, while in R everything is contains in the base R.
The main advantage of using vectors is to vectorize operation, instead of using loops to compute a function on each elements of a list, we can simply perform it on blocks of data in a more efficient may (faster, less memory used)


let's first see an example to understand th benefit of using vectorized operation :

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

import numpy as np
import time 

arr = np.arange(2000000)

lst = list(range(2000000))

t = time.time()
arr_2 = arr*2
time.time() - t

t = time.time()
lst_2 = [i*2  for i in lst]
time.time() - t


```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

vec = c(1:2000000)

t = Sys.time()
vec_2 = vec*2
Sys.time() - t

t = Sys.time()
vec_2_loop <- rep(0,length(vec))
for(i in 1:length(vec)){
  vec_2_loop[i] = vec[i]*2
    }
Sys.time() - t

```

</div>
</div>
</div>

# Dealing with Arrays

## Random array 


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

#generate an array
arr = np.random.randn(4,5)

arr

# vectorise calculation
arr+arr**2

# show dimension
arr.shape

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

#generate a matrix
mat <- matrix(runif(4*5),4,5)

mat

# vectorise calculation
mat+mat**2

# show dimension
dim(mat)

```

</div>
</div>
</div>

## Array-Oriented Programming

Imagine two square matrices representing interaction between entities. Each matrix is a basic network: having interaction = 1, no interaction = 0.
You want to know if two entities that are linked in one network are also linked in an other network. Instead of checking each entities interaction sequentially in a loop, one can simply use matrix to do element-wise calculation and get the same answer.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

#generate a list
list_1 = [[0,1,0,1],[1,0,1,0],[0,1,0,1],[1,0,1,0]]
network_1 = np.array(list_1)

#generate a sequence
list_2 = [[0,1,0,0],[1,0,0,0],[0,0,0,1],[0,0,1,0]]
network_2 = np.array(list_2)

network_12 = network_1*network_2
network_12
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

#generate a sequence
list_1 <- list(c(0,1,0,1),c(1,0,1,0),c(0,1,0,1),c(1,0,1,0))

network_1 <- do.call(rbind,list_1)

#generate a sequence
list_2 = list(c(0,1,0,0),c(1,0,0,0),c(0,0,0,1),c(0,0,1,0))

network_2 <- do.call(rbind,list_2)

network_12 = network_1*network_2
network_12

```

</div>
</div>
</div>

# Conditional Logic


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

network_12 == 1

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R
 
network_12 == 1

```

</div>
</div>
</div>


On thing that we often need is to modify specific value in a array that are satisfiying a condition. The first thing we want to know is which cells is satisfying the condition. In Python np.where is very common while in R we use the which function.

See how the execution time is changing depending on the size of the vector

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

arr = np.arange(200000)


t = time.time()
results = [(i if i%2==0 else 0) for i in arr]
time.time() - t


t = time.time()
results = np.where(arr%2!=0,0,arr)
time.time() - t
```

```{python}
# Python 

arr = np.arange(20000000)


t = time.time()
results = [(i if i%2==0 else 0) for i in arr]
time.time() - t


t = time.time()
results = np.where(arr%2!=0,0,arr)
time.time() - t
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

vec <- seq(1,200000)

t <- Sys.time()
results <- rep(0,length(vec))
for(i in 1:length(vec)){
  if(vec[i]%%2!=0){
    results[i] = 0
  } else {
    results[i] = vec[i]
  }
}
Sys.time() - t


t <- Sys.time()
results <- ifelse(vec%%2!=0,vec,0)
Sys.time() - t

```

```{r, collapse = TRUE}
# R

vec <- seq(1,20000000)

t <- Sys.time()
results <- rep(0,length(vec))
for(i in 1:length(vec)){
  if(vec[i]%%2!=0){
    results[i] = 0
  } else {
    results[i] = vec[i]
  }
}
Sys.time() - t


t <- Sys.time()
results <- ifelse(vec%%2!=0,vec,0)
Sys.time() - t


## other way using where

t <- Sys.time()
vec[which(vec%%2==0)] = 0
Sys.time() - t 

```

</div>
</div>
</div>

# Operation on array



<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

# nb of link
network_12.sum()/2
# same since there is only 0 and 1 
(network_12>0).sum()/2

# share of link 
network_12.mean()


# nb link by entities
network_12.sum(axis=1)

# share of link across entities (by columns)
network_12.mean(1)



```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

# nb of link
sum(network_12)
# same since there is only 0 and 1 
sum(network_12[which(network_12>0,arr.ind = T)]) # see the arr.ind to get the two coordinates

# share of link 
mean(network_12)


# nb link by entities
apply(network_12,MARGIN = 2,FUN = sum)

# share of link across entities (by columns)
apply(network_12,MARGIN = 2,FUN = mean)


```

</div>
</div>
</div>

- Reorder the object using sort()


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 

arr = np.random.randn(10)

arr
arr.sort()


```

```{python}
# Python 

arr = np.random.randn(4,3)

arr
arr.sort(1)


```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

arr <- runif(10)

arr
sort(arr)

```

```{r, collapse = TRUE}
# R

arr <- matrix(runif(4*3),4,3)

arr
apply(arr,MARGIN = 2, FUN = sort)

```

</div>
</div>
</div>

# Algebra


<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 
from numpy.linalg import inv, qr 
 
X = np.array([[0,1,5,1],[2,1,3,1],[2,1,9,6],[7,2,1,0],[8,3,5,5]]) 
Y = np.array([0,1,2,5,4])

X
X.T.dot(X) 
# same as 
XtX = np.dot(X.T,X)
# same as 
X.T @ X

# inverse Matrix
inv(XtX)

# Identity matrix
np.diag(np.ones(3))

```

```{python}
# Python 

XtY = np.dot(X.T,Y)

# OLS 
Beta = inv(XtX).dot(XtY)
Beta
```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R

 
X <- do.call(rbind,list(c(0,1,5,1),c(2,1,3,1),c(2,1,9,6),c(7,2,1,0),c(8,3,5,5))) 
Y <- c(0,1,2,5,4)
 
t(X)%*%X

XtX <- t(X)%*%X

# inverse Matrix
solve(XtX)

# Identity matrix
diag(3)
```

```{r, collapse = TRUE}
# R


XtY = t(X)%*%Y

# OLS 
Beta = solve(XtX) %*% XtY
Beta

```

</div>
</div>
</div>

# Sparse

When dealing with large array that are sparse (lot of element equalt to 0), it is inneficient to store this object as a dense matrix. Large matrix are sometimes difficult to store in the memory because they are weighty, when they dealing with sparse we can store it and do calculation with them in a very efficient way using the 'Compressed Sparse Column' (CSC) format. See how it change the weight of the matrix when changing the format to sparse matrix .

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python}
# Python 
 
from scipy.sparse import csr_matrix

X = np.zeros((10000,10000))

idx = list(np.ndindex(X.shape))[np.random.randint(1,10000**2,1000)]

X[idx] = 1
```

```{python}
# Python 

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r, collapse = TRUE}
# R
library(Matrix)

mat = matrix(0,10000,10000)
idx = round(runif(1000,1,10000**2))

mat[idx] = 1

print(object.size(mat),units = 'auto')

sparse_mat = Matrix(mat,sparse = T)
print(object.size(sparse_mat),units = 'auto')
```

```{r, collapse = TRUE}
# R

mat = matrix(0,10000,10000)
idx = round(runif(100000,1,10000**2))

mat[idx] = 1

print(object.size(mat),units = 'auto')

sparse_mat = Matrix(mat,sparse = T)
print(object.size(sparse_mat),units = 'auto')
```

</div>
</div>
</div>

