---
title: "exo_chap_2"
---


<style>
div.python pre { background-color: #d0f5a9; }
</style>

<style>
div.r pre { background-color: #dfdddf; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
``` 


<a href="3_Arrays_Vectors.html"> Go back to chapter 2</a>

# Exercice 1

The series, $1^{1} + 2^{2} + 3^{3} + ... + 10^{10} = 10405071317$.

Find the last ten digits of the series, $1^{1} + 2^{2} + 3^{3} + ... + 1000^{1000}$.
## R 

<div class = "r">
```{r}
# R

1000**1000

library(gmp)

sum_n_power_last_digit <- function(n, last_digit){
  sum_n = stringr::str_split(as.character(sum(as.bigz(seq(1, n))**(seq(1, n)))),'')[[1]]
  return(sum_n[(length(sum_n)-last_digit):length(sum_n)])
}

t = Sys.time()
sum_n_power_last_digit(1000,10)
Sys.time() - t

```

</div>
## Python
<div class = "python">
```{python}
# Python 
import numpy as np
import time

# default numpy dtype may not be appropriate to deal with very large number, python can do it, you can specify to numpy to use it as a python object

np.power(np.arange(1,15),np.arange(1,15))
np.power(np.arange(1,15, dtype="object"),np.arange(1,15, dtype="object"))

# Corentin
t = time.time()
print(str(sum( [i**i for i in list(range(1, 1001))] ))[-10:])
time.time() - t


# In a function with python object
def sum_n_power_last_digit(n,last_digit):
  sum_n = np.sum(np.power(np.arange(1,n+1, dtype="object"),np.arange(1,n+1, dtype="object")))
  return str(sum_n)[-last_digit:]

t = time.time()
sum_n_power_last_digit(1000,10)
time.time() - t
```

</div>


# Exercice 2


## Exercice 1 

If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.

### R
<div class = "r">
```{r,collapse=T}
get_sum_multiples_below1 <- function(below,multiple_1,multiple_2){
  sum <- 0
  for(i in 1:(below-1)){
    if((i %% multiple_1 == 0)|(i %% multiple_2 == 0)){
      sum <-  sum + i 
    }
  }
  return(sum)
}

t <- Sys.time()
get_sum_multiples_below1(10000,3,5)
Sys.time() - t



get_sum_multiples_below2 <- function(below,multiple_1,multiple_2){
  seq <- seq(1,(below-1))
  multiples <- sapply(seq,FUN = function(x){if((x %% multiple_1 == 0) | (x %% multiple_2 == 0)){x}else{0} })
  return(sum(multiples))
}

t <- Sys.time()
get_sum_multiples_below2(10000,3,5)
Sys.time() - t



get_sum_multiples_below3 <- function(below,multiple_1,multiple_2){
  seq <- seq(1,(below-1))
  multiples <- seq[which((seq %% multiple_1 == 0) | (seq %% multiple_2 == 0))]
  return(sum(multiples))
}

t <- Sys.time()
get_sum_multiples_below3(10000,3,5)
Sys.time() - t


```

</div>

### Python

<div class = "python">

```{python,collapse=T}
import time
import numpy as np

def get_sum_multiples_below1(below,mutiple_1,multiple_2):
  sum = 0
  for i in range(below):
    if (i % mutiple_1 == 0) or (i % multiple_2 == 0):
      sum+= i 
      
  return sum

t = time.time()
get_sum_multiples_below1(10000,3,5)
time.time() - t


def get_sum_multiples_below2(below,mutiple_1,multiple_2):
  multiples = [i if ((i % mutiple_1 == 0) or (i % multiple_2 == 0)) else 0 for i in range(below)]
  return sum(multiples)

t = time.time()
get_sum_multiples_below2(10000,3,5)
time.time() - t


def get_sum_multiples_below3(below,mutiple_1,multiple_2):
  seq = list(range(below))
  multiples = map(lambda i: i if ((i % mutiple_1 == 0) or (i % multiple_2 == 0)) else 0, seq) 
  return sum(multiples)

t = time.time()
get_sum_multiples_below3(10000,3,5)
time.time() - t



def get_sum_multiples_below4(below,mutiple_1,multiple_2):
  seq = np.arange(below)
  seq = seq[((seq % mutiple_1 == 0) | (seq % multiple_2 == 0))]
  return sum(seq)

t = time.time()
get_sum_multiples_below4(10000,3,5)
time.time() - t


# Mohammad

t = time.time()
div = sum(list(filter(lambda x: x % 3 == 0 or x % 5 == 0, range(1,1000))))
print(div)
time.time() - t

``` 

</div>

## Exercice 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

### R
<div class = "r">
```{r,collapse=T}


nth_prime <- function(nth){
  primes = c(1,2)
  i = 2
  while(length(primes)<nth){
    i <- i+1
    is_prime = TRUE
    for(j in 2:(round(sqrt(i))+1)){
      if(i%%j == 0){
        is_prime = FALSE
      }
      if(is_prime == FALSE){
        break
      }
    }
    if(is_prime == TRUE){
      primes = c(primes,i)
    }
  }
  
  return(primes[nth])
}

t <- Sys.time()
nth_prime(10001)
Sys.time() - t


nth_prime2 <- function(nth){
  primes = c(2,3)
  i = 3
  while(length(primes)<nth){
    i <- i+1
    ratio = i%%seq(2,round(sqrt(i))+1) == 0
    if(!any(ratio)){
      primes = c(primes,i)
    }
  }
  return(primes[nth])
}


t <- Sys.time()
nth_prime2(10001)
Sys.time() - t


```

</div>

### Python

<div class = "python">
```{python,collapse=T}
# Python 
# Astrid
from math import *



def est_premier(nbr):
  if nbr==1 :
    return False
  if nbr%2==0 :
    return False
  for i in range(3, int(sqrt(nbr)+1), 2):
    if nbr%i==0:
      return False
  return True



t = time.time()
p=1
count=1
while count !=10001 :  
  p+=2
  if est_premier(p):
    count+=1
print(p)

time.time() - t

# Corentin
import math

def return_ith_prime(ith):
    prime = 3
    for j in range(1, ith-1):
        for i in range(prime+2, prime*2, 2):
            if all(i % np.arange(3, int(math.sqrt(i)+1), 2) != 0):
                prime = i
                break
    return(prime)

t = time.time()
print(return_ith_prime(10001))
time.time() - t



def nth_prime2(nth):
  primes = [2,3]
  i = 3
  while len(primes)<nth: 
    i+=1
    ratio = np.mod(i,np.arange(2,int(np.sqrt(i)+1))) == 0
    if not any(ratio):
      primes.append(i)
    
  
  print(primes[nth-1])


t = time.time()
nth_prime2(10001)
time.time() - t
``` 

</div>

# Exercice 3

2520 is the smallest number that can be divided by each of the numbers from 1 to 10 without any remainder.

What is the smallest positive number that is evenly divisible by all of the numbers from 1 to 20?

## R
<div class = "r">
```{r}
# R

smallest_multiple <- function(n){
    i = 0
    results = 1
    exp = 5
    while(i<n){
        i = ifelse(i+9 < n,i+10,n)
        limit = 10**exp
        exp = exp +5
        
        seq_ = seq(i-9,i)
        integer = seq(results,limit,results)
        integer = matrix(rep(integer,length(seq_)),length(integer),length(seq_))
        divider = matrix(rep(seq_,dim(integer)[1]),dim(integer)[1],length(seq_),byrow = T)
        results = apply(integer%%divider,1,FUN = base::sum)==0 
        results = min(integer[results,])
        
    }
    return(results)
    }

t = Sys.time() 
smallest_multiple(20)
Sys.time() - t

```


</div>
## Python

<div class = "python">
```{python}
# Python
# Corentin

def smallest_multiple(condition_numbers = np.arange(1, 11)):
    for l in range(int(len(condition_numbers)/2)):
        condition_numbers = np.insert(np.delete(condition_numbers, np.where(condition_numbers[-1] % condition_numbers == 0)), 
                                      0, condition_numbers[-1])
#     return condition_numbers
    for i in range(2, len(condition_numbers)**10, 2):
        for j in condition_numbers:
            if i%j != 0:
                break
            if j == condition_numbers[-1]:
                return i
        
print(smallest_multiple(np.arange(1, 11)))

t = time.time()
print(smallest_multiple(np.arange(1, 21)))
time.time() - t



```

```{python}
# Python

import numpy as np
import time

t = time.time() 
def smallest_multiple(n):
    i = 0
    results = 1
    exp = 5
    while i<n:
        i = i+10 if (i+9) < n else n
        limit = 10**exp
        exp += 5
        
        seq = np.arange(i-9,i+1)
        integer = np.arange(results,limit,results)
        divider = np.ones([len(integer),1])*seq
        integer = np.array([integer]*len(seq)).T
        results = np.min(integer[np.where(np.fromiter(map(sum,np.mod(integer,divider)), dtype = int) == 0)])
    return results
    
time.time() - t
smallest_multiple(20)
t = time.time() 

```

```{python}
# Python
# Mohammad
import numpy as np
import pandas as pd
import time
import math
import time

def DivisibleParN(N):
    ##Calcul des nombres premiers qui sont inférieurs à N et les stocker dans jj
    global jj
    jj=[1]
    XX=[1]
    YY=[1]
    ZZ=[1]
    Di=np.arange(2,N+1)
    Seq2=np.arange(2,3)
    Seq3=np.arange(2,int(math.sqrt(Seq2[-1]))+1,1)
    n=0
    while jj[-1]<N:
          if all (Seq2[-1]%x!=0 for x in Seq3):
                jj=np.append(jj,Seq2[-1])
                Seq2=np.append(Seq2,Seq2[-1]+1)
                Seq3=np.arange(2,int(math.sqrt(Seq2[-1]))+1,1)
          else:
                Seq2=np.append(Seq2,Seq2[-1]+1)
                Seq3=np.arange(2,int(math.sqrt(Seq2[-1]))+1,1)
    jj=jj[0:(len(jj)-1)]

    #########Calcul du plus petit commun multiple
    for x in Di:
        for y in Di:
            if ((round(x ** (1 / y)) ** y == x)&(x ** (1 / y)in jj)):
                ZZ=np.append(ZZ,x ** (1 / y))
                XX=np.append(XX,x)
                YY=np.append(YY,y)
    V=pd.DataFrame(dict(XX=XX, YY=YY,ZZ=ZZ))
     
    # Trouver la factorisation où il se produit le plus.(ex: entre 4, 8 et 16 on on va garder que le 16)
    V_mod = V.drop_duplicates(subset=['ZZ'], keep='last')
    number_column = V_mod.loc[:,'XX']
    numbers = number_column.values
    for ii in jj:
        if all (x%ii!=0 for x in numbers): numbers = np.append(numbers,ii)
    return np.prod(numbers)
    
t = time.time()
print (DivisibleParN(20),'est divisible par tous les chiffres.')
time.time() - t

```


</div>


# Exercice 4
```{python}
np.sum(np.array(list(str(2**1000)),dtype=np.int))

# Nikola
print(sum(int(i) for i in str(2**1000)))

```

# Exercice 5

```{python}
# Python
# Tamara

import numpy as np
import time
 
t = time.time()

def isPalindrome(p):
    for i in range (p.size//2):
        if p[i]!=p[-i-1]:
            return False
    return True

def cutNb(n):
    count=10
    res=np.array([])
    while n//10 != 0:
        res=np.append(res,[n%10])
        n=n//10
    res=np.append(res,[n%10])
    return np.flip(res)  #inverse l'ordre de la liste

maxpal=0
maxi=0
maxj=100

for i in range(999, 100, -1):
    for j in range(i, maxj, -1): 
        if isPalindrome(cutNb(i*j)):
            if maxpal<i*j: 
                maxpal=i*j
                maxi=i
                maxj=j
            break
   
            

print('le palindrome est:',maxpal,'avec i=',maxi,'et j=',maxj)

time.time() - t   

# Julian

t = time.time()
pal = []
for i in range(100, 1000):
    for x in range(100, 1000):
        n = str(i*x)
        if n[0] == n[-1] and n[1] == n[-2] and n[2] == n[-3]:
            pal.append(i*x)


print(max(pal))
time.time() - t

#
t = time.time()
three_digits = np.arange(100,1000)
products = np.repeat(three_digits, len(three_digits))*np.tile(three_digits,len(three_digits))
palindromes = list(filter(lambda x: str(x)==str(x)[::-1],list(products)))
print(max(palindromes))
time.time() - t


```
<a href="3_Arrays_Vectors.html"> Go back to chapter 2</a>
