---
title: "Exercice Chapter 1"
output: html_document
---

<style>
div.python pre { background-color: #e9f7f5
; }
</style>

<style>
div.r pre { background-color: #fff5fd; }
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
``` 


<a href="2_Basics.html"> Go back to chapter 1</a>

# Exercice 1



If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.


## R 

<div class = "r">

```{r}
# R

get_sum_multiples_below1 <- function(below,multiple_1,multiple_2){
  sum <- 0
  for(i in 1:(below-1)){
    if((i %% multiple_1 == 0)|(i %% multiple_2 == 0)){
      sum <-  sum + i 
    }
  }
  return(sum)
}

t <- Sys.time()
get_sum_multiples_below1(10000,3,5)
Sys.time() - t
```

</div>

## Python

<div class = "python">

```{python}
# Python 

import time

def get_sum_multiples_below1(below,mutiple_1,multiple_2):
  sum = 0
  for i in range(below):
    if (i % mutiple_1 == 0) or (i % multiple_2 == 0):
      sum+= i 
      
  return sum

t = time.time()
get_sum_multiples_below1(10000,3,5)
time.time() - t


```


```{python}
# Python 

def get_sum_multiples_below2(below,mutiple_1,multiple_2):
  multiples = [i if ((i % mutiple_1 == 0) or (i % multiple_2 == 0)) else 0 for i in range(below)]
  return sum(multiples)

t = time.time()
get_sum_multiples_below2(10000,3,5)
time.time() - t

```


</div>


# Exercice 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

## R 

<div class = "r">


```{r}
# R

nth_prime <- function(nth){
  primes = c(1,2)
  i = 2
  while(length(primes)<nth){
    i <- i+1
    is_prime = TRUE
    for(j in 2:(round(sqrt(i))+1)){
      if(i%%j == 0){
        is_prime = FALSE
        break
      }
    }
    if(is_prime == TRUE){
      primes = c(primes,i)
    }
  }
  
  return(primes[nth])
}

t <- Sys.time()
nth_prime(10001)
Sys.time() - t
```
</div>

## Python

<div class = "python">

```{python}
# Python 
import numpy as np


def nth_prime(nth):
  primes = [1,2]
  i = 2
  while len(primes) < nth+1:
    i+=1 
    is_prime = True
    for j in range(2,(int(np.sqrt(i))+1)):
      if i%j == 0:
        is_prime = False
        break
        
    if is_prime == True:
      primes.append(i)
      
  print(primes[-1])


t = time.time()
nth_prime(10001)
time.time() - t



```

</div>


# Exercie 3

You are given the following information, but you may prefer to do some research for yourself.

- 1 Jan 1900 was a Monday.
- Thirty days has September,
    April, June and November.
    All the rest have thirty-one,
    Saving February alone,
    Which has twenty-eight, rain or shine.
    And on leap years, twenty-nine.
- A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

How many Sundays fell on the first of the month during the twentieth century (1 Jan 1901 to 31 Dec 2000)?

## R

<div class = "r">

```{r}

get_sundays <- function(year,first_sunday){
  
  # Get number of days per month
  if((year %% 100 != 0 & year%%4 == 0) | year %% 400 == 0){
    month_length =  c(31,29,31,30,31,30,31,31,30,31,30,31)
  } else {
    month_length = c(31,28,31,30,31,30,31,31,30,31,30,31)
  }
  
  # total number of days
  nb_days = sum(month_length)
  # position of the first days of the month
  cumsum_year = cumsum(month_length)-month_length+1
  
  # position of all sundays
  sundays = seq(first_sunday,nb_days,7)
  
  # get first sunday of the following year
  next_sunday_position = sundays[length(sundays)] - nb_days + 7
  
  nb_sundays_first = length(which(sundays %in% cumsum_year))
  
  return(c(next_sunday_position,nb_sundays_first))
}


t = Sys.time()
# Intialize with 1900, we know that the first sunday is the 7th.
year_result = get_sundays(1900,7)


# compute the sum for each year
nb_sundays = 0
for(i in seq(1901,2000)){
  next_sunday = year_result[1]
  year_result = get_sundays(i,next_sunday)
  nb_sundays = nb_sundays + year_result[2]
}
Sys.time() - t


# results
nb_sundays
``` 

</div>

## Python 

<div class = "python">

```{python}
# Python
import numpy as np 

def get_sundays(year,first_sunday):
  # check if there is a leap
  if (year % 100 != 0 and year%4 == 0) or year % 400 == 0 :
    month_length = [31,29,31,30,31,30,31,31,30,31,30,31]
  else:
    month_length = [31,28,31,30,31,30,31,31,30,31,30,31]
    
  # total number of days
  nb_days = sum(month_length)
  # position of the first days of the month
  cumsum_year = np.cumsum(month_length)-month_length+1
  
  # position of all sundays
  sundays = list(range(first_sunday,nb_days,7))
  
  # get first sunday of the following year
  next_sunday_position = sundays[-1] - nb_days + 7
  
  nb_sundays_first = len(np.where(np.isin(sundays,cumsum_year)==True)[0])
  
  return next_sunday_position, nb_sundays_first


t = time.time()
# Intialize with 1900, we know that the first sunday is the 7th.
year_result = get_sundays(1900,7)

# compute the sum for each year
nb_sundays = 0

for i in range(1901,2001):
  next_sunday = year_result[0]
  year_result = get_sundays(i,next_sunday)
  nb_sundays = nb_sundays + year_result[1]


time.time() - t
nb_sundays


``` 


</div>


<a href="2_Basics.html"> Go back to chapter 1</a>