---
title: "Exercice Chapter 1"
output: html_document
---

<style>
div.python pre { background-color: #d0f5a9; }
</style>

<style>
div.r pre { background-color: #dfdddf; }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE,collapse = T)
``` 


<a href="2_Basics.html"> Go back to chapter 1</a>

# Exercice 1



If we list all the natural numbers below 10 that are multiples of 3 or 5, we get 3, 5, 6 and 9. The sum of these multiples is 23.

Find the sum of all the multiples of 3 or 5 below 10000.


## R 

<div class = "r">

```{r}
# R

get_sum_multiples_below1 <- function(below,multiple_1,multiple_2){
  sum <- 0
  for(i in 1:(below-1)){
    if((i %% multiple_1 == 0)|(i %% multiple_2 == 0)){
      sum <-  sum + i 
    }
  }
  return(sum)
}

t <- Sys.time()
get_sum_multiples_below1(10000,3,5)
Sys.time() - t
```

</div>

## Python

<div class = "python">

```{python}
# Python 

import time

def get_sum_multiples_below1(below,mutiple_1,multiple_2):
  sum = 0
  for i in range(below):
    if (i % mutiple_1 == 0) or (i % multiple_2 == 0):
      sum+= i 
      
  return sum

t = time.time()
get_sum_multiples_below1(10000,3,5)
time.time() - t


```


```{python}
# Python 

def get_sum_multiples_below2(below,mutiple_1,multiple_2):
  multiples = [i if ((i % mutiple_1 == 0) or (i % multiple_2 == 0)) else 0 for i in range(below)]
  return sum(multiples)

t = time.time()
get_sum_multiples_below2(10000,3,5)
time.time() - t

```


```{python}
# Python 
# Aristote

t = time.time()
vec = [t for t in range(1,10001) if t%3==0 or t%5==0]
s = 0
for v in vec :
    s += v
print(s)
time.time() - t

```

</div>


# Exercice 2

By listing the first six prime numbers: 2, 3, 5, 7, 11, and 13, we can see that the 6th prime is 13.

What is the 10 001st prime number?

## R 

<div class = "r">


```{r}
# R

nth_prime <- function(nth){
  primes = c(1,2)
  i = 2
  while(length(primes)<nth){
    i <- i+1
    is_prime = TRUE
    for(j in 2:(round(sqrt(i))+1)){
      if(i%%j == 0){
        is_prime = FALSE
      }
      if(is_prime == FALSE){
        break
      }
    }
    if(is_prime == TRUE){
      primes = c(primes,i)
    }
  }
  
  return(primes[nth])
}

t <- Sys.time()
nth_prime(10001)
Sys.time() - t
```

```{r}
# R
# Mohammad

t = Sys.time()
Chiffre=c(1,2,3,4,5)
n=1
  while (n<10001){
    if (all(Chiffre[length(Chiffre)]%%Chiffre[3:length(Chiffre)-1]!=0)){
      n=n+1
      #print(Chiffre[length(Chiffre)])
    } 
    #if (n==10000) {break}
    val=Chiffre[length(Chiffre)]+1
    Chiffre<-append(Chiffre,val)}
print(c('10001st prime number is',Chiffre[length(Chiffre)]))
Sys.time() - t
```
</div>

## Python

<div class = "python">

```{python}
# Python 
import numpy as np


def nth_prime(nth):
  primes = [1,2]
  i = 2
  while len(primes) < nth:
    i+=1 
    is_prime = True
    for j in range(2,(int(np.sqrt(i))+1)):
      if i%j == 0:
        is_prime = False
      
      if is_prime == False:
        break
        
    if is_prime == True:
      primes.append(i)
      
  print(primes[nth-1])


t = time.time()
nth_prime(10001)
time.time() - t



```

</div>

<div class = "python">

```{python}
# Python 
# Tamara


from math import *

def isPrime(n):
    b=int(sqrt(n))
    for i in range(2,b):
        if n%i==0:
            return False
    return True


t = time.time()
i=1
p=1

while i<10001:
    p+=1
    if isPrime(p):
        i+=1    


time.time() - t



```

```{python}
# Python 
# Astrid


def est_premier(nbr):
  if nbr==1 :
    return False
  if nbr%2==0 :
    return False
  for i in range(3, int(sqrt(nbr)+1), 2):
    if nbr%i==0:
      return False
  return True



t = time.time()
p=1
count=1
while count !=10001 :  
  p+=2
  if est_premier(p):
    count+=1
print(p)

time.time() - t



```


</div>


# Exercie 3

You are given the following information, but you may prefer to do some research for yourself.

- 1 Jan 1900 was a Monday.
- Thirty days has September,
    April, June and November.
    All the rest have thirty-one,
    Saving February alone,
    Which has twenty-eight, rain or shine.
    And on leap years, twenty-nine.
- A leap year occurs on any year evenly divisible by 4, but not on a century unless it is divisible by 400.

How many Sundays fell on the first of the month during the twentieth century (1 Jan 1901 to 31 Dec 2000)?

## R

<div class = "r">

```{r}
nb_sundays_y <- function(year,sunday_first){
  # check if there is a leap
  if(year %% 4 != 0 | year == 1900){
    month_length <-  c(31,28,31,30,31,30,31,31,30,31,30,31)
  } else {
    month_length <-  c(31,29,31,30,31,30,31,31,30,31,30,31)
  }
  
  # count the first sunday given if equal to 1
  nb_sunday_first = ifelse(sunday_first == 1,1,0)  

  # set the first sunday date
  next_sunday = sunday_first
  m <- 0
  for(i in month_length){
    m <- m+1
    # compute all sundays of the given month
    while(next_sunday <= i){
      next_sunday = next_sunday + 7
    }
    # setting first sunday date for the next month
    next_sunday = next_sunday - i
    
    # add to count if next month start with a sunday 
    # and if this is still the current year
    if(next_sunday == 1 & m!=12){
      nb_sunday_first = nb_sunday_first + 1
    }
  }
  return(c(nb_sunday_first,next_sunday))
}



t <- Sys.time()
# Intialize with 1900 for which we know that the first sunday is the 6th.
year_results <- nb_sundays_y(1900,6)
nb_sunday <- c() 

# compute the sum for each year
for(year in 1901:2000){
  year_results <- nb_sundays_y(year,year_results[2])
  nb_sunday <- c(nb_sunday,year_results[1]) 
}
  
# results
sum(nb_sunday)

Sys.time() - t
``` 

```{r}
# R
# Mohammad

t = Sys.time()

date=c()
Y=c(1900:2000)
for(year in Y){
  if((year-1900)%%4 != 0){
    date<- append(c(1:31, 1:28, 1:31, 1:30, 1:31, 1:30, 1:31, 1:31, 1:30, 1:31, 1:30, 1:31),date)
  }else {
    date<-append(c(1:31, 1:28, 1:31, 1:30, 1:31, 1:30, 1:31, 1:31, 1:30, 1:31, 1:30, 1:31),date)
  }}
ND <- sum(which(date[365:length(date)] == 1) %% 7 == 0)
print(c('Nombre de dimanche coincide avec le prmier jour du mois',ND))


Sys.time() - t
``` 

</div>

## Python 

<div class = "python">

```{python}
# Python

def nb_sundays_y(year,sunday_first):
  # check if there is a leap
  if year % 4 != 0 or year == 1900:
    month_length = [31,28,31,30,31,30,31,31,30,31,30,31]
  else:
    month_length = [31,29,31,30,31,30,31,31,30,31,30,31]
    
  # count the first sunday given if equal to 1
  nb_sunday_first = 1 if sunday_first == 1 else 0  

  # set the first sunday date
  next_sunday = sunday_first
  m = 0
  for i in month_length:
    m += 1
    # compute all sundays of the given month
    while next_sunday <= i:
      next_sunday += 7
    
    # setting first sunday date for the next month
    next_sunday = next_sunday - i
    
    # add to count if next month start with a sunday 
    # and if this is still the current year
    if next_sunday == 1 and m!=12:
      nb_sunday_first += 1
    
  
  return(nb_sunday_first,next_sunday)


t = time.time()
# Intialize with 1900 for which we know that the first sunday is the 6th.
year_results = nb_sundays_y(1900,6)
nb_sunday = [] 

# compute the sum for each year
for year in range(1901,2001):
  year_results = nb_sundays_y(year,year_results[1])
  nb_sunday.append(year_results[0]) 
# results
sum(nb_sunday)

time.time() - t



``` 

```{python}
# Python 
# Corentin 
import pandas as pd

t = time.time()
# Create a DF of years~months

dic = {}
for y in range(1900, 2001):
    dic[y] = range(1, 13)

calendar = pd.DataFrame(dic)

# Find the day of each month's first day
for year in calendar:
    for month in range(12):
        # 30 days months
        if month in [4, 6, 9, 11]:
            days_in_month = 30
        # leap years february
        elif (month == 2) and ((year%4 == 0 and year%100 > 0) or (year%400 == 0)):
            days_in_month = 29
        # february
        elif (month == 2):
            days_in_month = 28
        # others have 31 days
        else:
            days_in_month = 31
            
        # fill each cell with the day
        # all months except january
        if (year != 1900 or (year == 1900 and month != 0)) and (month > 0):    
            calendar.iloc[month, year-1900] = calendar.iloc[month-1, year-1900] + days_in_month%7
        # January
        elif (year != 1900 or (year == 1900 and month != 0)) and (month == 0):
            calendar.iloc[month, year-1900] = calendar.iloc[11, year-1901] + days_in_month%7
        # normalize on the 7-day week
        if calendar.iloc[month, year-1900] > 7:
            calendar.iloc[month, year-1900] -= 7
    
## 1st day Sunday in th 20th

# remove 1900
twentieth = calendar.iloc[:, 1:]

# function : calaculates the number of 1st day of month sundays in a given year
def sundays(vec):
    return(sum(vec==7))

# sum over all years from the dataframe
answer = sum(twentieth.apply(sundays))
answer

time.time() - t

``` 


</div>


<a href="2_Basics.html"> Go back to chapter 1</a>