---
title: "Dplyr"
---


<style>
div.python pre { background-color: #e9f7f5
; }
</style>

<style>
div.r pre { background-color: #fff5fd; }
</style>





```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,error=TRUE, warnings = F)
library(reticulate)
```


<a href="5_Data_Analysis_R.html">Go back to chapter 4</a> 

# Import datasets


<div class = "r">

```{r}
# R dplyr
library(tidyverse)
library(tictoc)
# Import the two datasets using readr's read_csv funciton
tic()
table_1 = read_csv(file = 'data/table_1.csv')
table_2 = read_csv(file = 'data/table_2.csv')
cities = read_csv(file = 'data/cities_loc.csv')
toc()
```

</div>

# Clean

Remove articles with no pmid and no DOI, and all articles before 1975

<div class = "r">

```{r}
# R dplyr

tic()
table_1 = table_1 %>%
  filter(pmid != 'null' & doi != 'null')

table_2 = table_2 %>%
  filter(pmid != 'null' & year_pub > 1975)
toc()

```

</div>

# Merge

Merge the two datasets, pmid is unique for each paper


<div class = "r">

```{r}
# R dplyr
tic()
final_table = inner_join(table_1, table_2, by = 'pmid')
toc()

final_table = na.omit(final_table)
final_table
```

</div>



# Number of authors

Create a new variable with the number of authors for each paper



<div class = "r">

```{r}
# R dplyr

tic()
final_table = final_table %>% 
  rowwise() %>%
  mutate(nb_aut = dim(str_match_all(authors, '<AUTHOR>')[[1]])[1]) 
toc()

final_table = final_table %>% 
  filter(nb_aut > 0)

final_table %>% 
  select(authors, nb_aut)
```

</div>


#  plot distribution

plot distribution for the log(number of authors +1)


<div class = "r">

```{r}
# R 

plot(density(log(final_table$nb_aut+1)))
```

</div>



# Search for DL/ML/NN in title

 How many papers contains 'deep learning' or 'machine learning' and 'neural network' (also with a 's' for neural networks) in their title ? Create a binary variable to save this information. What is the mean of authors for ML papers and non-ML papers ? Transform has_data and oa into binary variable also, what is the share of ML paper that are oa ?


<div class = "r">

```{r}
# R dplyr

tic()
final_table = final_table %>%
  mutate(title = tolower(title),
         ML = ifelse(str_detect(title,'deep learning|machine learning|neural networks?'), 1, 0),
         has_data = ifelse(has_data == 'Y', 1, 0),
         oa = ifelse(oa == 'Y', 1, 0))
toc()


sum(final_table$ML)
sum(final_table$oa)
sum(final_table$has_data)

tic()
final_table %>% 
  group_by(ML) %>%
  summarize(mean_nb_aut = mean(nb_aut))
toc()

tic()
final_table %>% 
  group_by(ML) %>% 
  summarize(n = sum(oa)/n(),
            citation = mean(cited))
toc()
```

</div>

# Clean up pub_type

 Clean up pub_type, for simplicity just get the first type



<div class = "r">

```{r}
# R dplyr
tic()
final_table = final_table %>%
  mutate(pub_type = str_match_all(pubtype, '\\[\\"(.*?)\\"')[[1]][2])
toc()

final_table %>% 
  select(pubtype,pub_type)

```

</div>


# Highest mean/sd of citation

What is the pub type with the highest mean/sd of citation ? (use cited and the cleaned pub_type)


<div class = "r">

```{r}
# R dplyr

final_table %>% 
  group_by(pub_type) %>% 
  summarize(mean = mean(cited),
            sd = sd(cited)) %>%
  arrange(desc(mean))

final_table %>%
  group_by(pub_type) %>% 
  summarize(mean = mean(cited),
            sd = sd(cited)) %>%
  arrange(desc(sd))

```

</div>

# Most representative countres by year ?

 Which are the most representative countres by year ? Store this information in an other tibble by keep only pmid and authors, get the country for each author from the loc_cities.csv. You may want to separate rows for each authors to get all countries involved in the paper,if an authors have multiple affiliations, take the first one. 

<div class = "r">

```{r}
# R dplyr

countries = tolower(unique(cities$country))
tic()
countries_tibble = final_table %>%
  select(pmid,authors) %>% 
  separate_rows(authors, sep = '<AUTHOR>')
toc()

countries_tibble %>% 
  select(authors)

tic()
countries_tibble = countries_tibble %>%
  rowwise() %>% 
  filter(authors !="" & !str_detect(authors, '<AFFILIATION>None'))
toc()

countries_tibble %>% 
  select(authors)

tic()
countries_tibble = countries_tibble %>%
  group_by(pmid,authors) %>%
  mutate(authors = strsplit(authors,'<AFFILIATION>')[[1]][2]) %>% 
  filter(!duplicated(pmid, authors))
toc()

countries_tibble %>% 
  select(authors)
```

</div>



<div class = "r">

```{r,warning=F}
# R dplyr
tic()
countries_tibble  = countries_tibble %>%
  mutate(authors = str_replace_all(authors, ' USA', 'United States'),
         authors = str_replace_all(authors, ' UK', 'United Kingdom'),
         authors = str_replace_all(authors, ' Korea', 'South Korea'),
         authors = tolower(authors)) 
toc()

tic()
countries_tibble = countries_tibble %>%
  group_by(pmid, authors) %>%
  mutate(country = ifelse(any(str_detect(authors, countries)),
                          countries[str_detect(authors, countries)][1],
                          "")) %>%
  select(-authors)
toc()

tic()
countries_tibble = countries_tibble %>%
  filter(country != '') %>%
  distinct(pmid, country)
toc()
countries_tibble


tic()
countries_tibble = left_join(countries_tibble,
                             final_table[,c('pmid','year_pub')],
                             by = 'pmid')
toc()

tic()
countries_pub = countries_tibble %>% 
  group_by(country, year_pub) %>%
  summarize(n = length(unique(pmid))) %>% 
  arrange(desc(n))
toc()
```

</div>

#  Country bar plot with plotly

 Select the top 25 of countries involved in coronavirus research since 2001, plot the evolution on a bar chart with plot_ly


Two usefull links:

- https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html


- https://plotly.com/r/bar-charts/

<div class = "r">

```{r, warning= F}
# R 
library(plotly)
library(kableExtra)

plot_top_n  = function(n_top,year){
  
  countries_top = countries_pub %>%
    filter(year_pub >= year) %>%
    group_by(country) %>% 
    summarize(n = sum(n)) %>% 
    arrange(desc(n)) %>% 
    ungroup() %>% 
    slice(1:n_top)

  countries_top %>% 
    kbl(caption = "Top 25 most active countries in coronavirus research  (2001-2021)") %>%
    kable_classic(full_width = F, html_font = "Cambria")
  
  countries_pub %>% 
    filter(country %in% as.matrix(countries_top$country) & year_pub >= year) %>% 
    plot_ly(x = ~year_pub, y = ~n, color = ~country) %>% 
    layout(yaxis = list(title = 'Count',type = 'log'), barmode = 'stack')
}

plot_top_n(25,2001)

```

</div>

