
## Regular Expressions

Regex are a way to express how a character sequence should be matched in a more flexible way than what we are used to do with Ctrl+F.
Regular expression are commonly used in text manipulation, usually such patterns are used by string-searching algorithms for "find" or "find and replace" operations on strings, or for input validation. 

Except for control characters, (+ ? . * ^ $ ( ) [ ] { } | \), all characters match themselves. You can escape a control character by preceding it with a backslash.

Il y a principalement deux normes :

• PCRE (Perl-Compatible Regular Expressions). Associé au langage de
programmation PERL, elle est notamment exploitée par la bibliothèque re de
Python.
• POSIX étendu, fruit d’un effort de normalisation, elle est censée être plus
simple, mais est en revanche plus lente. L’intérêt pour nous est que R s’appuie
sur cette norme (sauf à indiquer explicitement l’option perl=TRUE).
Regular Expression Examples

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>

Character classes

Example|Description
-|-:
[Pp]ython|Match "Python" or "python"
rub[ye]|Match "ruby" or "rube"
[aeiou]|Match any one lowercase vowel
[0-9]|Match any digit; same as [0123456789]
[a-z]|Match any lowercase ASCII letter
[A-Z]|Match any uppercase ASCII letter
[a-zA-Z0-9]|Match any of the above
[^aeiou]|Match anything other than a lowercase vowel
[^0-9]|Match anything other than a digit

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>
Special Character Classes

Example|Description
-|-:
.|Match any character except newline
\d|Match a digit: [0-9]
\D|Match a nondigit: [^0-9]
\s|Match a whitespace character: [ \t\r\n\f]
\S|Match nonwhitespace: [^ \t\r\n\f]
\w|Match a single word character: [A-Za-z0-9_]
\W|Match a nonword character: [^A-Za-z0-9_]

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>

Repetition Cases

Example|Description
-|-:
ruby?|Match "rub" or "ruby": the y is optional
ruby*|Match "rub" plus 0 or more ys
ruby+|Match "rub" plus 1 or more ys
\d{3}|Match exactly 3 digits
\d{3,}|Match 3 or more digits
\d{3,5}|Match 3, 4, or 5 digits

Nongreedy repetition
This matches the smallest number of repetitions −

Example|Description
-|-:	
<.*>|Greedy repetition: matches "<python>perl>"
<.*?>|Nongreedy: matches "<python>" in "<python>perl>"

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>

Grouping with Parentheses

Example|Description
-|-:
\D\d+|No group: + repeats \d
(\D\d)+|Grouped: + repeats \D\d pair
([Pp]ython(, )?)+|Match "Python", "Python, python, python", etc.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>

Backreferences
This matches a previously matched group again −

Example|Description
-|-:
([Pp])ython&\1ails|Match python&pails or Python&Pails
(['"])[^\1]*\1|Single or double-quoted string. \1 matches whatever the 1st group matched. \2 matches whatever the 2nd group matched, etc.

Alternatives
Sr.No.	Example & Description
	
python|perl|Match "python" or "perl"
rub(y|le))|Match "ruby" or "ruble"
Python(!+|\?)|"Python" followed by one or more ! or one ?

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>
Anchors
This needs to specify match position.

Example|Description
-|-:
^Python|Match "Python" at the start of a string or internal line
Python$|Match "Python" at the end of a string or line
\APython|Match "Python" at the start of a string
Python\Z|Match "Python" at the end of a string
\bPython\b|Match "Python" at a word boundary
\brub\B|\B is nonword boundary: match "rub" in "rube" and "ruby" but not alone
Python(?=!)|Match "Python", if followed by an exclamation point.
Python(?!!)|Match "Python", if not followed by an exclamation point.

<div class = "row">
<div class = "col-md-6">
<div class = "python">

```{python,echo = T,collapse = TRUE}
# Python 
import re

```

</div>
</div>
<div class = "col-md-6">
<div class = "r">

```{r,echo=T,collapse = TRUE}
# R
library(stringr)

```

</div>
</div>
</div>