---
title: "Basics"
author: "Pierre Pelletier"
date: "3 septembre 2020"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)
```

# Install R-studio and Python

## R-studio

## Python

# Operations

## Arithmetic Operations
Python and R can both work as classical calculator, using "+", "-", "*" and "/" you can do arithmetic operations in both language.
```{python,echo = T,collapse = TRUE}
# Python 

1+2
1-2
1/2
1*2

```


```{r,echo=T,collapse = TRUE}
# R

1+2
1-2
1/2
1*2
```

You can also apply exponantiation, Modulo and floor division easily n both language. Please note that in R you can writte the exponantiation "^". You cannot do that in Python but both use "**" for exponantiation. Use this one.

```{python,echo = T,collapse = TRUE}
# Python 

2**8 # exponantiation
2^8 == 2**8 # False
8%3 # modulo
8//3 # floor division
```


```{r,echo=T,collapse = TRUE}
# R

2**8 # exponantiation
2^8 == 2**8 # TRUE
8%%3 # modulo
8%/%3 # floor division

```


##  Comparison Operators

In order to compare some values we can use comparison operators to investigate in a given value is equal to, not equal, greater than ... etc

Theses operators are the same in bth language.


```{python,echo = T,collapse = TRUE}
# Python 

2==8
2!=8
2<8
2>8
2<=8

```


```{r,echo=T,collapse = TRUE}
# R

2==8
2!=8
2<8
2>8
2<=8
```