---
title: "Scheduling exercise solution"
author: "Pierre Pelletier"
date: "2/7/2021"
output:
  pdf_document: 
    toc: true
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      error=FALSE,
                      collaps = TRUE)

```

\newpage

# Exercice 

There are XX groups in the class, the orals must be spread over 6 hours, in two 3-hour sessions.
For each group the duration of the oral is about ~20 min + ~10 min for Q&A.
We will want to take two 15-minute breaks in the after 1h30 of courses. 
Write a program that randomly assigns a schedule to each group, write out this list of names and schedules in a table. Have a look on this : https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html


# Examples

## Using Python
```{python}

import pandas as pd

test_py = pd.DataFrame(
    {
     'schedule':['h1','h2','h3'],
     'name':['A','B','C']
     }
    )

``` 


## Using R

```{r}
library(dplyr)

test_R = data.frame(
  'schedule'=c('h1','h2','h3'),
  'name'=c('A','B','C')
  )

``` 

# Get Python object in R

```{r}
library(reticulate)

py$test_py
``` 

# Create latex tables

```{r}
library(kableExtra)

py$test_py %>% 
  kbl(booktabs = T,
      caption = "Schedules using pandas' DF") %>%
    kable_classic(full_width = F, html_font = "Cambria")

test_R %>% 
  kbl(booktabs = T,
      caption = "Schedules R's DF") %>%
    kable_classic(full_width = F, html_font = "Cambria")
``` 
- then we put ```echo = FALSE,``` in ```` ```{r, echo = FALSE}``` ```` and we don't see the input anymore


```{r, echo = F}
test_R = data.frame(
  'schedule'=c('h1','h2','h3'),
  'name'=c('A','B','C')
  )

py$test_py %>% 
  kbl(booktabs = T,
      caption = "Schedules using pandas' DF") %>%
    kable_classic(full_width = F, html_font = "Cambria")

test_R %>% 
  kbl(booktabs = T,
      caption = "Schedules R's DF") %>%
    kable_classic(full_width = F, html_font = "Cambria")
``` 

