---
title: "Basics"
author: "Pierre Pelletier"
date: "3 septembre 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(reticulate)

#  md_document:
#    variant: markdown_github
```

# Operations

## Arithmetic Operations
Python and R can both work as classical calculator, using "+", "-", "*" and "/" you can do arithmetic operations in both language.
```{python,echo = T,collapse = TRUE}
# Python 

1+2
1-2
1/2
1*2

```


```{r,echo=T,collapse = TRUE}
# R

1+2
1-2
1/2
1*2
```

You can also apply exponantiation, Modulo and floor division easily in both language. Please note that in R you can writte the exponantiation "^". You cannot do that in Python (it is used for antoher operation) but both use "**" for exponantiation. Use this one for clarity.

```{python,echo = T,collapse = TRUE}
# Python 

2**8 # exponantiation
2^8 == 2**8 # False
8%3 # modulo
8//3 # floor division
```


```{r,echo=T,collapse = TRUE}
# R

2**8 # exponantiation
2^8 == 2**8 # TRUE
8%%3 # modulo
8%/%3 # floor division

```


##  Comparison Operators

In order to compare some values we can use comparison operators to investigate in a given value is equal to, not equal, greater than ... etc

Theses operators are the same in both language.


```{python,echo = T,collapse = TRUE}
# Python 

2==8
2!=8
2<8
2>8
2<=8

```


```{r,echo=T,collapse = TRUE}
# R

2==8
2!=8
2<8
2>8
2<=8
```


# Variables

ref used : https://www.practicaldatascience.org/html/vars_v_objects.html

Variables are used to store data. In Python and R, we do not need to declare a variable before assigning a value to this variable. You can think of a variable as a name to refers to an object. There is a big difference between python and R on the way the computer is storing object and variables that we will see in the end of the section.



## textual and numerical variables
In R textual data are called 'character', in Python it is a string, abbreviate as 'str'. " or ' are used when you set a textual variable in both laguage, but you can also set a variable to textual variable if needed.

Numerical variables a decomposed in three types. In Python you'll have integer, float and complex. In R float is called 'double', reffering to double precision floating point. Without entering into details it means that the number have a precision up to 15 decimals. Float can refers to a precision of 7 decimals, however in Python the data-type 'float' has a precision of 15 numbers. They are ways to get more precisions using Numpy for example. 

Finaly you can check easily the data-type of the variable that you are dealing by asking the language with the command 'type' in Python, or 'typeof' in R.

```{python,echo = T,collapse = TRUE}
# Python 
import sys 
sys.float_info.dig # number of decimals

a = 1
type(a)

b = 1.1
type(b)

c = 1.1+2j
type(c)

```


```{r,echo=T,collapse = TRUE}
# R

a = 1
typeof(a)
a_int = 1L
typeof(a_int)

b = 1.1
typeof(b)

c = 1.1+2i
typeof(c)

```


## Other Data-Type
refused : https://thomas-cokelaer.info/tutorials/python/lists.html

There is a lot of datatype in both language that we will discover along the course, for now we will focus only on basic data-type that are directly available without using anyother package or library. In Python the most common are List, Turple, sets, Dictionnary. While in R you will have Vector, Matrix, List, Data frame and factor.

## Lists 
append() and extend() methods
```{python,echo = T,collapse = TRUE}

a = [1, 2, 3]
a.count(2) # count elements of the list which are exactly equal to 2

a.sort(reverse = True)
a
# access the element of a list
a[0]
a.index(3)
a[1:]
a[:1]
a[0:-1]
a[:]

# modify
b = [0, 0, 0]
a.append(b)
a
a.extend([4, 5, 6])
a
a.insert(2,[1,2])
a
a.remove(b)
a

# Operators on lists

a += [7,8] # works as extend
a = a*2 # replicate the list n times
a

```


```{r,echo=T,collapse = TRUE}
# R

a = 1
typeof(a)
a_int = 1L
typeof(a_int)

b = 1.1
typeof(b)

c = 1.1+2i
typeof(c)

```




## Objects and Variables, difference between R and Python /!\


In R, a variable and an object are the same things, they refers to the same entities. If you assign to a new variable a variable that already exist, it will refers to two different objects.

Things in Python are more conventional, but since some people learned R before Python it's important to notice that variables refers to an objet but they are still two entities. It means that two variables can refer to a single object. If two variables are pointing the same object, the modification made with one variable will be also available when using the other variable.There some 
See the example above for Python. In R you would need to modify variable setting an new object to this variable using '=' o '<-'.
However not all data-type of Python are muttable.



```{python,echo = T,collapse = TRUE}
# Python 
import copy
##mutable
a = [1,6,1] #list

b = a
b_copy = copy.copy(a[:])
b_copy = copy.copy(a)
b.append(8)
print(a)



```

# control flow

<div class = "row">
<div class = "col-md-6">
```{python,echo = T,collapse = TRUE}
# Python 
evens = []
for i in range(10):
    if i % 2 == 0:
        evens.append(i)
evens
# beter to writte this
[i for i in range(10) if i % 2 == 0]
```
</div>
<div class = "col-md-6">
```{r, warning = FALSE, echo=T}
lala = 1+1
lala
```
</div>
</div>
# Dealing with strings 

https://thomas-cokelaer.info/tutorials/python/strings.html

